---
title: "LECTURE 05: Estimation"
subtitle: "ENVS475: Exp. Design and Analysis"
author: "<br/><br/><br/>Spring 2023"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: true
---

```{r setup, echo = FALSE, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE, fig.retina = 2)

library(ggplot2) 

set.seed(2052)
wolf_x <- round(rnorm(8, 38, 4))
wolf_x2 <- round(rnorm(8, 38, 4))
mean_1 <- round(mean(wolf_x), 1)
mean_2 <- round(mean(wolf_x2), 1)
sd_1 <- round(sd(wolf_x), 1)
sd_2 <- round(sd(wolf_x2), 1)
sem_1 <- round(sd_1 / sqrt(length(wolf_x)), 1)
sem_2 <- round(sd_2 / sqrt(length(wolf_x2)), 1)  
```

class: inverse

# Outline

<br/>
#### 1) Descriptive Statistics

<br/>  

#### 2) Inferential Statistics

<br/> 

#### 3) Confidence Intervals

<br/> 

---
# Review: populations vs samples

#### Population  
- A collection of subjects of interest  

- Often, a biologically meaningful unit  

- Sometimes a process of interest  

#### Sample

- A finite subset of the population of interest, i.e. the data we collect  

- Samples allow us to draw inferences about the population  

- Good samples are:
    + Random  
    + Representative  
    + Sufficiently large  

---

# Review: parameters vs statistics

### Parameters 
- Attributes of the population  
  + Mean ( $\mu$ )  
  + Variance ( $\sigma^2$ )  
  + Standard deviation ( $\sigma$ )  
- Parameters are the quantities of interest, and usually unknown  

--
### Statistics
- Attributes of the sample  
  + Mean ( $\bar{y}$ or $\hat{\mu}$ )  
  + Variance ( $s^2$ or $\hat{\sigma}^2$ )  
  + Standard deviation ( $s$ or $\hat{\sigma}$ )  
- Often treated as estimates of parameters
---

class: inverse, center, middle
# Descriptive Statistics

---
# Descriptive Statistics

### Measures of central tendency

- **Sample mean**

$$\large \bar{y} = \frac{\sum_{i=1}^n y_i}{n}$$

<br/>

--
- **Median**: "middle" of the data  

  + 50% of data is below, 50% of data is above median  
  
  + Useful for non-normal, or skewed data  
  
  + If data is truly normal, mean = median  
  

<br/>
--

- **Mode**: Most frequent observation in the data

---

### Central Tendency Example:

* Weight of Wolves (kg) from Yellowstone:
`r round(wolf_x)`

<br/>

* Mean (rounded): `r round(mean(wolf_x),1)`  
  * `mean()`

<br/>

* Median (rounded): `r round(median(wolf_x),1)`  
  * `median()`  

<br/>

* Mode: 35  
  * no default in function in R  
  * For small data sets, `sort()` vector and count repeats
  * `r sort(round(wolf_x))`  

---

### Measures of dispersion

- Sample variance  
  - How far, on average, are the *observations* from the mean?  
  - squared scale - different from data

$$\large s^2 = \frac{\sum_{i=1}^n (y_i - \bar{y})^2}{n-1}$$

<br/>

--
- Sample standard deviation  
  - How far, on average, are the *observations* from the mean?  
  - same scale as data

$$\large s = \sqrt{s^2}$$

<br/>

--
- Range  
  * min and max values of data

---
### Dispersion Example:

* Weight of Wolves (kg) from Yellowstone:
`r round(wolf_x)`

<br/>

* Variance: `r round(var(wolf_x),1)`  
  * `var()`  
  * Remember, different scale!

<br/>

* Standard Deviation: `r round(sd(wolf_x),1)`  
  * `sd()`  
  * same scale as data  
  * SD is mostly what we will use  

<br/>

* Range: `r round(range(wolf_x))`  
  * `range()`  
---
# Describing the sample  

* When describing the data, always include the mean and a measure of dispersion  

  + be sure to label measure of dispersion (var or SD?)
  
> The weight of Yellowstone wolves is `r round(mean(wolf_x), 1)` $\pm$ `r round(sd(wolf_x),1)` kg (mean $\pm$ SD)  

* Alternate:  

> The average ( $\pm$ SD)  weight of wolves in Yellowstone is `r round(mean(wolf_x), 1)` $\pm$ `r round(sd(wolf_x),1)` kg 

---
# Inferential Statistics

* So far, we've been discussing *descriptive statistics* for our observations  

* Analyses often interested in comparing means, or estimates of $\mu$.  

* This requires *inferential statistics*  
---
# Inferential Statistics

* Weight of Wolves (kg) from Yellowstone:
`r round(wolf_x)`
* Mean: `r round(mean(wolf_x),1)`  
* SD: `r round(sd(wolf_x),1)`  

<br/>
### How confident are we in this value?  
* If we measured another 8 wolves, what would our estimates be?  

* 2nd Sample: `r round(wolf_x2)`
* 2nd Mean: `r round(mean(wolf_x2), 1)`  
* 2nd SD: `r round(sd(wolf_x2), 1)`  

---

# Inferential Statistics
### Standard error of the mean: SEM  

<br/>

* How far, on average, is the sample mean $\bar{y}$ from the true mean $\mu$?  

<br/>

* We could repeat experiment over and over again to get a *sampling distribution*  

<br/>

  * Rarely done due to time, logistics, funding, etc.  
  
<br/>

* Luckily, we can estimate the SEM from a single sample

---

# Inferential Statistics
### Standard error of the mean: SEM  

$$\Large SEM = \frac{s}{\sqrt{n}}$$
<br/>

* Remember, SEM tells us how far, on average, the sample mean $\bar{y}$ is from the true mean $\mu$  

* SEM is like a standard deviation *of the mean*


---
### Helpful(?) mnemonic 

<br/>

* SD: dispersion for the data (**D**)  

  * How far is an individual observation from the sample mean?  

<br/>

* SEM: dispersion for the mean (**M**)  

  * How far is the sample mean from the true mean?  

<br/>

  * Note that SEM is sometimes just abbreviated as SE, so this mnemonic isn't always obvious. 
  
---
### Point estimates  

* $\bar{y}$ is a point estimate of the true mean, $\mu$  
* point estimates on their own are of limited value  

* Always include a measure of precision when you report a mean

$$\Large \bar{y} \pm SEM$$

--

#### Wolf sample 1:  
* Mean: `r mean_1`
* SEM: `r sem_1`  

--

$35.8 \pm 0.8$ (mean $\pm$ SEM)

--

* Note similarity with sample description previously  

--

  * Always label your dispersion estimate! 

--

* You can also present the interval by completing the equation:

* Interval = `r c(mean_1 - sem_1, mean_1 + sem_1)`

---

class: inverse, center, middle
# Confidence Intervals

---
# Confidence Intervals  
- Confidence Intervals (CI) is a range of estimates for an unknown parameter.  
- CI computed for different confidence levels: most often 95%  
- Assuming a normal distribution, 95% CI is approximately:

$$\Large 95\%CI =\bar{y} \pm 2*SEM$$  

  + Recall empirical rule: 95% of data within $\pm 2~SD$  
  
  + In future classes we will discuss equation for CI from a *t*-distribution, i.e., modify value of 2 in above equation    

- If we calculated a 95% CI from $n$ samples, about 95% of them would contain the true population mean  

  + **NOTE** it *does not mean* that we are 95% sure that the true mean is in the CI  
  
---

### Repeated samples from Population  
- 100 repeated samples of wolf weights  
- Increased to 25 wolves per sample  

```{r CI-plot,  echo=FALSE}
library(dplyr)
set.seed(4251)
true_mean = 38
true_sd = 4

samp <- data.frame(Experiment = 1:100,
                   time = 1:100,
                   Estimate = numeric(length = 100),
                   SE = numeric(length = 100))

for(i in 1:100){
  x <- rnorm(25, true_mean, true_sd)
  samp$Estimate[i] = mean(x)
  samp$SD[i] = sd(x)
  samp$SE = sd(x)/sqrt(25)
}

samp <- dplyr::mutate(samp,
                      LCI = Estimate - 1.96 * SE,
                      UCI = Estimate + 1.96 * SE,
                      Coverage = 
                        ifelse(LCI <=true_mean & UCI >= true_mean, "Yes", "No"))

ggplot(samp,
       aes(x = Experiment,
           y = Estimate)) +
  geom_errorbar(
    aes(ymin = LCI,
        ymax = UCI,
        color = Coverage),
    width = 0) +
  geom_point(aes(color = Coverage)) +
  scale_color_manual(values = c("red", "black")) +
  geom_hline(yintercept = true_mean,
             color = "red",
             linetype = "longdash") +
  scale_y_continuous(name = "",
                     limits = c(33, 44),
                     breaks = seq(from = 33, to = 44)) +
  scale_x_continuous(limits = c(0, 100)) +
  theme_classic() +
  geom_segment(
    aes(y = 43, yend = 43,
        x = 1, xend = 12), 
               color = "red", linetype = "longdash") +
  annotate(
    x = 15,
    y = 43,
    label = "True parameter value",
    geom = "text",
    hjust = 0) +
  geom_point(
    aes(y = 42,
        x = 6)) +
  annotate(x = 15,
           y = 42,
           label = "Sample mean and 95% confidence interval",
           geom = "text",
           hjust = 0) 
```

* 97 / 100 CIs contain true mean ~97%  
---
### Approximate 95% CI for Yellowstone Wolves  

$$\Large 95\%CI =\bar{y} \pm 2*SEM$$
--
<br/>

$$ \large \bar{y} = `r mean_1`$$  
--
<br/>

$$ \large SEM = `r sem_1` $$
--
<br/>

$$\large 2 * SEM = `r 2* sem_1`$$

--
<br/>

$$\Large 95\%CI = `r c(mean_1-2*sem_1, mean_1+2*sem_1)`$$   

---

### Using Confidence Intervals to test hypotheses  

* Previous research has indicated that mean wolf weight in the Arctic is 33 kg.  

* Does this differ from the wolves in Yellowstone?  

* CI for Yellowstone wolf weight (1st sample)  

$$95\%CI =\bar{y} \pm 2*SEM$$  

<br/>

$$35.6 \pm 2 * 0.8 = (34,~37.2)$$

* Arctic weight of 33 is outside (below) 95% CI for Yellowstone wolves.  

* **Interpretation**: The data support the conclusion, with 95% confidence, that the weight of Yellowstone Wolves is between 34 and 37.2 kg, and this is higher than the mean estimated weight for wolves in the Arctic of 33 kg. 

---
### Changing the Confidence Interval  

* We can approximate a 99% CI by multiplying the SEM by 3

$$99\%CI =\bar{y} \pm 3*SEM$$ 

$$35.6 \pm 3 * 0.8 = (33.2,~38)$$

* How does the 99% CI compare with the 95% CI?  
  + *i.e.*, which is wider and narrower?  
  + Why do you think this is?  
  
---

### Using Confidence Intervals to test hypotheses  

* Data from British Columbia indicates that the mean weight of wolves there is 34.2 kg.  

--

* Do the weight of wolves in BC differ from those in Yellowstone?  

--

* 34.2 falls within the 95% CI (34, 37.2) for Yellowstone Wolves.

--

* **Interpretation**: The data support the conclusion that the weight of Yellowstone Wolves is between 34 and 37.2 kg (95% CI), and that this is not different from the mean estimated weight of 34.2 for wolves in British Columbia.  

--

* Note we did not use CI's for Arctic or BC wolves.  
  
  * Using CI's is one form of (rough) hypothesis testing  
  
  * We can more formally test these hypotheses with t-test and general linear models  
  
  * We will get to this later in the course.  

---


# Looking ahead

### **Wednesday**: Point estimates and CI Lab

<br/>

### **Friday**: Hw 05 - For a grade

<br/>

### **Reading**: Hector Chapter 5

