---
title: "HW Template"
author: "PUT YOUR NAME HERE"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Make sure to put your name in the `author` spot at the top of the document and delete this text**

## Preliminaries

Put all `library()` calls here. Also load **ALL** data objects necessary for the home work assingment here. 

For example, the code below loads the `tidyverse` and the `lterdatasampler` packages. It also reads in the `galapagos-finches.csv` data file (make sure it is downloaded *and* in your `data/` folder). It also loads the `and_vertebrates` data which comes in the `lterdatasampler` package. 
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(lterdatasampler)
finch <- read_csv("data/galapago-finches.csv")
data(and_vertebrates)
```

## Problem 1 

All problems need a level 2 heading (i.e., two \#\#  ). 

All sub-problems need a level 3 heading (i.e., \#\#\#) as well as the number of the problem and sub-problem. 

For example:  

### Problem 1.1  

Below the level 3 heading, you can type plain text if needed to answer the question. If you need to run calculations in R, you will need to insert a new code chunk (mac: Cmd + Option + I, Windows Ctrl + Alt + I). 

```{r}
head(finch)
```

**Example answer for 1.1:** The first few rows of `finches` are shown. All variables are class double, except for `species` which is a character vector. 

### Problem 1.2  

```{r}
and_vertebrates %>%
  na.omit() %>%
  ggplot(aes(x = length_1_mm,
             y = weight_g,
             color = species)) +
  geom_point() +
  theme_dark() +
  scale_color_manual(values = c("hotpink", "cadetblue1"))
```

**Example answer for 1.2:** The slope for coastal giant salamander (pink) is steeper than cutthroat trout (blue). 

## Problem 2  

### 2.1  

## Problem 3  

### End of homework  