---
title: "ANOVA transformation problem"
subtitle: "Not in a homework yet"
date: "Written Spring, 2023"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
   warning = FALSE, message = FALSE,
   eval = FALSE,
   fig.align = 'center', fig.retina = 2,
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(ggplot2)
crop <- read.csv(here::here("data/clear_cut_factorial.csv"))

```

# problem 1: `clear_cut` 

### Load the data into R
```{r}
clear <- read.csv("data/clear_cut_factorial.csv")
```

This is a simulated data set looking at the effects that clear cuts at different elevations have on the number of bird species present in plots throughout the season. 

You should always print out the `names()` and `head()` of data objects when you load them into your R environment.

You may also want to look at the distinct/unique levels of each of the categorical/factor variables. 


1.1 Make a box plot the response variable (`spp`) separated by both grouping variables. 

1.2 Interpret the box plots here. Does it appear that there is equal variance across the groups? Why or why not?

1.3 Make a QQ-plot to assess the assumptions of a normal distirbution of residuals. 

1.4 Interpret the QQ-plot. Do the residuals look normally distributed? Why or why not?

1.5 Write out the hypotheses for a two factor ANOVA without interactions.  

  * **Hint** You should have two sets of null and alternative hypotheses.  
  
1.6 Fit a linear model using the `lm()` function, and save it as a new data object.  

1.7 Assess the last assumption of equal variance of residuals by making a plot of the fitted vs. residual values.  

1.8 Interpret the fitted vs. residual plot. Do the residuals appear to have equal variation? Why or why not?  

1.9 Regardless of your answers above, make a new variable which log-transforms your response variable, and save it as a new data object.  

  * **Hint** The formula for a log transformation will go inside of a `mutate()` function and should look something like this: `new_variable_name = log(response_variable + 1)`

1.10 Fit a linear model using the transformed data.  

  * **Hint** you will need to modify the model fit performed in problem 1.6 by changing the data object and the response variable.  
  
### Redo plots with transformed data  

1.11 Make a box plot the response variable (`spp`) separated by both grouping variables. 

1.12 Interpret the box plots here. Does it appear that there is equal variance across the groups? Why or why not?

1.13 Make a QQ-plot to assess the assumptions of a normal distirbution of residuals. 

1.14 Interpret the QQ-plot. Do the residuals look normally distributed? Why or why not?


1.15 Assess the assumption of equal variance of residuals by making a plot of the fitted vs. residual values.  

1.16 Interpret the fitted vs. residual plot. Do the residuals appear to have equal variation? Why or why not?  


1.17 Print out an anova table of the model using the transformed data (i.e., the second model that you fit)  

1.18 Write out the Statistical interpretation of the ANOVA table. Make sure you use the correct statistics (df's, F- and P-values) and relate it back to your null and alternative hypotheses.   

1.19 Perform a multiple comparison using Tukey's HSD test, and print out the results.  

1.20 Make a plot of the TukeyHSD multiple comparisons. 

1.21 Add a comment which has your interpretation of the Tukey HSD plot.  

  * **Note** Because we are using transformed data, you would need to back-transform the values for the estimated difference and 95%CI reported in the TukeyHSD output. We are going to skip that step for now, so here I'm just asking you to interpret the plots. The interpretation is the same whether data are transformed or not, i.e., does the CI contain zero or not?  
