---
title: "ENVS475 Homework 00: Setup"
subtitle: "Complete for a grade"
date: "`r format(Sys.time(), '%B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
   warning = FALSE, message = FALSE,
   fig.align = 'center', fig.retina = 2,
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
data("starwars")
```

# Complete this assignment for a grade.

### Learning Objectives

> Following this assignment students should:
>
> - Have R and Rstudio downloaded
> - Be able to access and open Rstudio
> - Have an R-Project for the course
> - Be able to download packages and load them into your working session

1. Download R and Rstudio
2. Make a new Project called "ENVS475" 
3. In the console, enter the following command and run it:

```{r, eval=FALSE}
install.packages("tidyverse")
```

**NOTE:** Do *NOT* put `install.packages()` commands in your scripts. Only ever run them in the console. 

* You can also install packages by clicking `Tools` --> `Install Packages` and then entering the package name in the text box and clicking `Install`  


4. Make a new script called "lastname_first-assignment.R" and save it in your `homework` folder in your ENVS396 Project. 

5. In your script, type in the following code and make sure it runs correctly:

```{r, eval=FALSE}
library(tidyverse)
data("starwars")
starwars %>% 
    filter(homeworld == "Tatooine") %>%
    select(name) %>%
    unique()
```


You should see the following output in your console:

```{r echo=FALSE}
library(tidyverse)
data("starwars")
starwars %>% 
    filter(homeworld == "Tatooine") %>%
    select(name) %>%
    unique()
```


## Assignment submission 

**For full credit, you must show your instructor that you have accomplished all of the above tasks.**