---
title: "Homework: dplyr + ggplot"
subtitle: "ENVS475: Experimental Analysis and Design"
date: "Spring 2024"
output: 
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
   warning = FALSE, message = FALSE,
   eval = FALSE,
   fig.align = 'center', fig.retina = 2,
  collapse = TRUE,
  comment = "#>"
)
finch <- read.csv(here::here("data/galapago-finches.csv"))
```

# Assignment Format

* Open a new R script
* Save your R script in the `homework/` folder
* Name your script according to: `full_name_hw##.R`
  + For example: `justin_pomeranz_hw01.R`
* Comment before each problem, and each sub-problem
* Make sure your results print out on `Source with echo`
  + If you're answer is saved in an object, make sure you print out that object afterwards
  + See that `volume` is on the last line of the example below

```{r assignment format, eval = FALSE}
# 2.1 example
w = 2
h = 3
l = 1.5
w
h
l
#2.2 example 
volume = w * h * l
volume
```

# Setup

## Load packages

* Make sure that `tidyverse` is downloaded on the machine you're using  

  + You can check this by clicking on the `Packages` tab in the *Files, Plots, Packages...* panel and scrolling down  
  
  + If you need to download the package, copy the following code into your Console panel and run it **Do not put it in your homework script*  
  
  + `install.packages("ggplot2")`   
  
* Copy the following code into the beginning of your homework script to load the `ggplot2` package:  

```{r, eval=FALSE}
# load libraries
library(tidyverse)
```

## Data

Download the `galapagos-finches.csv` data from D2L and put it in your `data/` folder in your Class R project. 

Read the data file into R using the following command:

```{r, eval=FALSE}
# read in full data set
finch <- read.csv("data/galapago-finches.csv")
```

# Exercises: `dplyr` and `ggplot2`

## Problem 0: get to know the data  
0.1 Use `head()` to print out the first few rows of the `finch` data.  

0.2 Print out the *last* few rows of the `finch` data. 

0.3 Print out the names of the columns in the `finch` data.

0.4 Use `select()` to print out only the `species` column.  

0.5 Use `select()` to print out the `year`, `species` and `blength` columns, in that order. 

0.6 Use `arrange()` to print out the `finch` data in order based on the `blength` column.  

0.7 Use `arrange()` to print out the `finch` data in order based on the `species`, `year`, and `bdepth` columns, in that order.  

0.8 Use `mutate()` to add a new column called `beak_area`, and have its value equal to `blength * bdepth`.  

0.9 Use `summarize` to calculate the average `blength` across the whole `finch` data set.  

0.10 Use `group_by()` and  `summarize` to calculate the average `blength` for each `species` in `finch` data set.  


## Problem: 1 Plot one continuous variable  
1.1 Make a histogram of the `blength` column.  

1.2 Make a histogram of the `blength` column and give it a descriptive title (5-7 words) and change the the x-axis label to be more easily read and include the units (mm).  

1.3 Make a histogram of the `blength` column with 100 bins.  

1.4 Make a histogram of the `blength` column where the binwidth is 1.  

1.5 Make a boxplot of the `blength`.  

1.6 Make a boxplot of the `blength` but change the fill color to one of your choice. (See a [list of color names](https://derekogle.com/NCGraphing/resources/colors) available in ggplot)  


## Problem 2: Plot with grouping variables  

2.1 Make a boxplot of the `blength`, but have the fill color vary based on the species.  

2.2 Make a boxplot of the `blength`, but have the fill color vary based on the year collected. (Careful of continuous vs. categorical variables. )  

2.3 Make a boxplot of the `blength`, but have the fill color vary based on the year, and have species mapped to the y-axis.    


## Problem 3: Plot two continuous variables  

3.1 Use `geom_point()` to make a scatter plot with `blength` on the x-axis, and `bdepth` on the y-axis.  

3.2 Use `geom_point()` to make a scatter plot with `blength` on the x-axis, and `bdepth` on the y-axis. Also include a linear regression on the plot for the whole data set  (there should be one line on your graph).  

3.3 Use `geom_point()` to make a scatter plot with `blength` on the x-axis, and `bdepth` on the y-axis. Also have the color of the points relate to the species, and include a linear regression on the plot for each species separately (there should be 2 lines on your graph).  

3.4 Repeat the graph you made in 3.3 above but include a separate facet for each year of data. 


## Problem 4: dplyr + ggplot

4.1 Use `mutate()` to create a new variable in the `finch` data set. This new variable should be equal to beak length $\times$ beak depth. Make a plain histogram showing the distribution of your new variable.  

4.2 Do the same thing as in 4.1, but this time, only plot the data from 1975, and map the fill color to the species variable.   

4.3 Do the same thing as in 4.2, but add a descriptive title and x-axis label.  

4.4 Do the same thing as in 4.1, but this time have the fill color mapped to species, and have a facet for each year.  