---
title: "ENVS475 Homework 5: point estimates"
subtitle: "Complete for a grade"
date: "Spring, 2023"
output: 
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
   warning = FALSE, message = FALSE,
   eval = FALSE,
   fig.align = 'center', fig.retina = 2,
  collapse = TRUE,
  comment = "#>"
)
```

# Complete this assignment for a grade.

# Assignment Format

* Open a new R script
* Save your R script in the `homework/` folder
* Name your script according to: `full_name_hw##.R`
  + For example: `justin_pomeranz_hw01.R`
* Comment before each problem, and each sub-problem
* Make sure your results print out on `Source with echo`
  + If you're answer is saved in an object, make sure you print out that object afterwards
  
# Setup

## Load packages

* Make sure that the `dplyr` and `ggplot2` packages are downloaded on the machine you're using
  + You can check this by clicking on the `Packages` tab in the *Files, Plots, Packages...* panel and scrolling down
  + If you need to download the package, copy the following code into your Console panel and run it **Do not put it in your homework script*
  + `install.packages("dplyr")` 
* Copy the following code into the beginning of your homework script to load the `dplyr` package:
```{r, eval=FALSE}
# load libraries
library(dplyr)
library(ggplot2)
```

## Data

`beak <- read.csv("data/g-fortis-beak-1975.csv")`

# Exercises Homework 05: point estimates

## Problem 1

## Problem 2

## Problem 3

## Problem 4

## Problem 5 Challenge (optional)

