---
title: "ENVS475 Homework 11: Factorial"
subtitle: "Not Graded"
date: "Spring, 2023"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
   warning = FALSE, message = FALSE,
   eval = FALSE,
   fig.align = 'center', fig.retina = 2,
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)

```

# This assignment will Not be graded.

# Assignment Format

* Open a new R script
* Save your R script in the `homework/` folder
* Name your script according to: `full_name_hw##.R`
  + For example: `justin_pomeranz_hw11.R`
* Comment before each problem, and each sub-problem
* Make sure your results print out on `Source with echo`
  + If you're answer is saved in an object, make sure you print out that object afterwards
  
# Setup

## Load packages

```{r, eval=FALSE}
# load libraries
library(dplyr)
library(ggplot2)
```

# Data

For this homework, you will need to download three .csv files from D2L: `moth_data.csv`, and `latin_square_crops.csv`.  

# Problem 1: `moth` 

### Load the data into R  
```{r}
moth <- read.csv("data/moth_data.csv")
```

This is simulated data looking at the effects of different pesticides for controlling the abundance of a moth pest species. The experiment was blocked based on different regions in the SE US. 

You should always print out the `names()` and `head()` of data objects when you load them into your R environment.

You may also want to look at the distinct/unique levels of each of the categorical/factor variables. 

### Wrangle the data

You may have noticed that the `Region` variable is a number. If we plug the data as-is into a linear model, R will treat this as a continuous variable. Since this represents different groups, we need to change the class of this variable to be a factor. 

Run the following code **before** answering the problems below. 

```{r}
moth <- moth %>%
  mutate(Region = factor(Region))
```

1.1 Make a box plot of the response variable (`caterpillar`) grouped with the herbicide treatment.  

  * **Note** do **not** include `Region` as a grouping variable. There is only one observation for each treatment in each region, and this is not enough to make a box plot.  

1.2 Interpret the box plots here. Does it appear that there is equal variance across the different treatments? Why or why not?  

1.3 Now, make a box plot of the response variable grouped by region (ignore the treatment variable)  

1.4 Interpret the box plots here. Does it appear that there is equal variance across the different regions? Why or why not?  

1.5 Fit a linear model using the `lm()` function, and save it as a new data object. Make sure to include both the `Treatment` and `Region` variables as main effects in the model.  

1.6 Write out the null and alternative hypotheses for this model.  


1.7 Print out the ANOVA table for the linear model you fit in problem 2.5.  

1.8 Write out your statistical interpretation of the ANOVA table. Make sure to report all appropriate statistics. 

1.9 Write out an interpretation of these results using plain words.  

1.10 To see how accounting for blocking effects can affect our model, fit a linear model without the region factor, and save it as a new data object.  


1.11 Print out the ANOVA table for your second model (without region)  

1.12 Write out the statistical interpretation of this second model. 


1.13 What does this mean in your own words? What happened?  


# Problem 2: `latin_square_crops`

### Read in the data

```{r}
latin_ag <- read.csv("data/latin_square_crops.csv")
```

This is a data set investigating the impact of tillage, fertilizer, and seed variety on yield (measured in cwt_year) in an agricultural setting. 

* `cwt_year` = this is the response variable (yield) measured in hundred-weight. (Hundred weight is a measurement in commodities trading. Basically, a higher number means more yield)

You should always print out the `names()` and `head()` of data objects when you load them into your R environment.

You may also want to look at the distinct/unique levels of each of the categorical/factor variables. 

### Experimental design

Here is an illustration of how the experimental field was set up. 

```{r, echo=FALSE}
latin_ag <- read.csv(here::here("data/latin_square_crops.csv"))
ls_mat <- matrix(latin_ag$seed, 5, 5)
row.names(ls_mat) <- unique(latin_ag$fertil)
colnames(ls_mat) <- unique(latin_ag$treat)
ls_mat
```

Variables: 
  * `treat` = each "column" in the field received a different tillage method  
  * `fertil` = each "row" in the field received a different fertilizer mix  
  * `seed` = "LETTER" = each "cell" in the field received a different seed variety  
  
You can see that each column and row has exactly one observation of each seed variety (A-E).   

2.1 Write out the null and alternative hypotheses for this experiemntal design.  

  * **Hint** You should have one set of hypotheses for each factor: Fertilizer, Tillage, and Seed.  
  
2.2 Fit a linear model for this analysis, and save it as a new data object.  

2.3 Print out the ANOVA table for your linear model.  

2.4 Write out the statistical interpretation for the main effects tested in this model. Be sure to include all appropriate statistics and relate it back to your null and alternative hypotheses.  

2.5 Perform a Tukey HSD multiple comparison for this model.   

2.6 Interpret the output of the Tukey HSD here.  

  * **Note** There are lots of pairwise comparisons here. To simplify,  you can just list the pairs within each factor which were significantly different.  

2.7 Make a plot of the Tukey HSD results.  

### Coda  

This concludes the homework assignment.  
