---
title: "Lab 08: t-tests"
subtitle: "ENVS475: Experimental Analysis and Design"
date: "Spring, 2023"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
bt_samp <- read.csv(here::here("data/body_temp_hr.csv"))
caterpillar <- read.csv(here::here("data/caterpillar.csv"))
```

## About

This lab will explore formal hypothesis testing by considering classic t-tests from a linear modelling perspective.  

## Preliminaries 

Recall that the general structure of a t-test is as follows:

$$\large t = \frac{difference}{SE}$$

### Packages  

For this lab, we will be using the `dplyr` and `ggplot2` packages. Make sure both packages are installed on your computer, and then run the following code:

```{r, eval=FALSE}
library(dplyr)
library(ggplot2)
```


### Data  

For the first two parts of this lab, we will be creating data sets in R. For the third part, you will need to download the `caterpillar.csv` from D2L and place it in your `data/` folder.  

# One sample t-test: Stream pH  

You may expect that streams around Grand Junction should have an approximately neutral pH of 7. Let's look at some simulated data and formally tests whether or not the average stream pH is significantly different from 7. Here, the value of interest is $\mu_0 = 7$  

## Formal Hypotheses  

We can state our hypotheses formally using Null and Alternatives as follows:

$$H_0: \mu = 7$$
$$H_A: \mu \ne 7$$

Since our Alternative hypothesis is $\ne$ this is a 2-tailed test. In other words we have to account for the fact that pH could be higher or lower than our null value of 7.  


We can express this as a simple linear model:  

$$y_i = \mu + \epsilon_i$$
where $y_i$ is a single observation of stream pH, $\mu$ is the population mean, and $\epsilon_i$ is the deviation from the expected value. Further, we expect the deviations to be distributed normally:

$$\epsilon \sim N(0, \sigma)$$

## Analysis: Manual   

Let's look at some **simulated data**, and load it into R as a vector. 

```{r}
# simulated data values
y <- c(6.2, 6.8, 7.3, 6.4, 6.3)
```

First, let's plot the data. For this example, we will use `geom_boxplot()`. Becasue `ggplot()` requires the data object to be a data.frame, we will quickly convert our vector `y` inside of the function. We will also add a dashed line to indicate where our value of interest (7) is.  

```{r}
ggplot(data = data.frame(y = y), 
       aes(y = y)) +
  geom_boxplot() +
  geom_hline(yintercept = 7,
             size = 1.5,
             linetype = "dashed",
             color = "pink") +
  theme_bw()
```



For the one-sample t-test, we will calculate all necessary variables manually, and then compare it to using the `lm()` function.  

Recall that we calculate a one-sample t-statistic with the following:

$$t_{statistic} =\frac{\bar{y} - \mu_{0} }{SEM_y}$$

Where $\bar{y}$ is the mean, $\mu_0$ is the value of interest (pH = 7) and $SEM$ is calculated as $s / \sqrt(n)$. Let's calculate these values in R and store them as objects. 

```{r}
y_bar <- mean(y)
y_bar
mu_0 <- 7
mu_0
y_sd <- sd(y)
y_sd
y_sem <- y_sd / sqrt(length(y))
y_sem
```

Now let's calculate the t-statistic

```{r}
t_stat <- (y_bar - mu_0) / y_sem
t_stat
```

Now we need to calculate the t-critical value for this t-distribution. We can use the `qt()` function. This is similar to the `qnorm()` function we used previously for the normal distribution, but now we are looking at the t-distribution. for `qt()`, we need to supply the quantile (0.025) and the degrees of freedom (df = n - 1 = 4). The standard $\alpha$ value is 0.05, but since this is a 2-tailed test we have to divide it by 2 (hence 0.025) in order to account for both tails. 

```{r}
t_crit <- qt(0.025, df = 4)
round(t_crit, 2)
```

Is our `t_stat` value $\ge$ our critical value? Note that we are going to use the absolute values so that we do not need to worry about using the proper > or < with negative numbers. 

```{r}
abs(t_stat) > abs(t_crit)
```

It is not, so we would fail to reject our null hypothesis. We can also calculate a p-value for our `t_critical` value by using the `pt()` function. We first calculate the probability of observing out `t_stat`, and then multiply it by 2 since this is a 2-tailed test. Also, we are going to use the absolute value of our `t_stat` and set the `lower.tail = FALSE` argument within `pt()`. This ensures that we don't need to worry about mixing up our + and - signs when calculating a p-value.  

```{r}
one_samp_p_val <- pt(abs(t_stat), df = 4, lower.tail = FALSE) * 2
one_samp_p_val
```

## Interpretation  

Based on our data, we estimate that the mean ($\pm$ SE) pH in streams near Grand Junction is 6.6 ($\pm$ `r round(y_sem, 2)`), and we fail to reject the null hypothesis that stream pH is equal to 7 ($t_{statistic} = -1.96, df = 4, p = 0.12$). 

## Analysis: using `lm()`

We can use the `lm()` function to automatically calculate our values and test the hypothesis formally. 

Recall that `y` is our observations of data, and our null value is 7. When we run a one-sample t-test in `lm()`, we have to subtract the value of interest from our data. Review the numerator in the formula for the one-sample t-statistic above. 

```{r}
one_sample_fit <- lm(y - mu_0 ~ 1)
summary(one_sample_fit)
```

We can also reprint the values we calculated by hand and locate them in the output table (note that we are subtracting mu_0 from y_bar to put it on the same scale as our `(Intercept)` estimate in the table):

```{r}
y_bar - mu_0
y_sd
y_sem
t_crit
one_samp_p_val
```

### **You should now be able to complete problem 1 in the homework assignment**

# Two-sample t-tests: Tree Densities  

Is the density of trees at high elevations different from low elevations?

$$H_0: \mu_{high} = \mu_{low}$$
$$H_A: \mu_{high} \ne \mu_{low}$$
## Data  

We will make a data.frame and call it `tree`. Normally, you should type out everything in this class for practice, but copy and paste the following code in order to avoid typing a mistake. 

```{r}
tree <- data.frame(
  elevation = rep(c("Low", "High"), each = 10),
  dens = c(16, 14, 18, 17, 29, 31, 14, 16, 22, 15, 2, 11, 6, 8, 0, 3, 19, 1, 6, 5))
```

Here, `dens` is the density of trees, and was simulated for this exercise. Let's once again plot our data to visualize how they are distributed. 

```{r}
ggplot(tree,
       aes(x = elevation, 
           y = dens,
           fill = elevation))+
  geom_boxplot() +
  theme_bw()
```


## Analysis: Manual  

Calculate statistics with dplyr. Recall that for a two-sample t-test we need the:

$$\large t = \frac{\bar{y}_1 -\bar{y}_2}{SEDM}$$
Where $\bar{y}_1$ and $\bar{y}_2$ are the different experimental groups, and SEDM is the standard error of the difference and is calculated as:

$$\large SEDM = \sqrt{SEM_1^2 + SEM^2_2}$$

Where subscript 1 and 2 refer to the different experimental groups, and that the $SEM$ for the $i$^th^ group is calculated as:

$$\large SEM_i = \frac{s_i}{\sqrt(n_i)}$$

Where $s$ and $n$ are the standard deviation and sample size for the $i$^th^ group, respectively. 

Where $i$ represents each group or sample.  

To summarize, we need to group the data by `elevation` groups (using `group_by()`) and then use a `summary()` call with `mean()`, `sd()` and `sqrt(n())` functions inside of it. Remember to name your "new" summary variables and use an `=` inside the summary call. After summarizing, we will use a `mutate()` function to add a new column/variable for the SEM. Finally, I will also modify the data.frame into a tibble to make the display in future calls easier to read.   

```{r}
tree_summary <- tree %>%
  group_by(elevation) %>%
  summarise(mean_density = mean(dens),
            sd_density = sd(dens),
            sqrt_n = sqrt(n())) %>%
  mutate(sem = sd_density / sqrt_n) %>%
  # change it from a tibble to a data.frame to make
  # next steps easier
  as.data.frame()
tree_summary
```


Unfortunately there is not an easy way to calculate the t-statistic using `dplyr`, so we will need to extract the variables individually (when we use the `lm()` all of these steps are automated, but we will continue to calculate it manually so you can see the process). Likewise, there is not a simple solution to extract individual variables from a `data.frame` using `dplyr`, so we will use the `data_object[row_number,column_number]` syntax to subset the values into new variables. This is also why we converted `tree_summary` into a `data.frame` in the previous step. 

If we want to extract the mean values, what numbers should we place in the square brackets in the following code? 

```
tree_summary[?,?]
tree_summary[?,?]
```

The following code will extract the means and SEMs from the `tree_summary` object. 

```{r}
y_high <- tree_summary[1,2]
y_high
y_low <- tree_summary[2,2]
y_low
sem_high <- tree_summary[1,5]
sem_high
sem_low <- tree_summary[2,5]
sem_low
```

Now, we will calculate the $t_{statistic}$ using this equation:

$$\large t = \frac{\bar{y}_{low} -\bar{y}_{high}}{\sqrt{SEM_{low}^2 + SEM^2_{high}}}$$


```{r}
t_stat_trees <- (y_low - y_high) / sqrt(sem_high^2 + sem_low^2)
t_stat_trees
```

Now that we have a t-statistic, we need to calculate our t-critical value to compare it with and decide if we can formally reject or fail to reject our null hypothesis. To calculate the t-critical value, we will once again use `qt()` funciton. Our degrees of freedom here is 18 (total n = 20, but we need to subtract 1 for each group: 20 - (1 + 1) = 18).  

```{r}
t_crit_trees <- qt(0.025, df = 18)
t_crit_trees
```

To compare our t-critical and t-statistic, we will once again use the absolute values to avoid mixing up our signs. 

```{r}
abs(t_stat_trees) > abs(t_crit_trees)
```

```{r}
two_samp_p_val <- pt(abs(t_stat_trees), df = 18, lower.tail = FALSE) * 2
two_samp_p_val
```
 
When you have a really low p-value like this, it's best to round instead of reporting the full value. Reporting the exact value adds a false sense of specificity. The three levels of significance are generally p < 0.05, p < 0.01, and p < 0.001. Here we would just say that the p-value is < 0.001. 
 
### Interpretation  

Based on the data, we estimate that the mean ($\pm$ SE) tree density at high and low elevations is `r y_high` ($\pm$ `r round(sem_high,2)`) and  `r y_low` ($\pm$ `r round(sem_low,2)`), respectively. Likewise, we can reject the null hypothesis and accept the alternative that tree densities differ based on elevation ($t_{statistic} = 4.97, df = 18, p < 0.001$). 

## 2-sample t-test using `lm()`  

The hand calculations for a 2-sample t-test are tedious. Luckily we can run the same analysis in just a few lines of code using the `lm()` function. 

```{r}
two_sample_fit <- lm(dens ~ elevation, data = tree)
summary(two_sample_fit)
```

## Interpreting `lm()` output  
The summary of the `lm()` has a lot of information. What we are primarily focused on here is the `elevationLow` row. The Estimate represents the *difference* between the two elevation groups. (Since "Low" is in the name, we can deduce that the `(Intercept)` is the estimated value for the other elevation group, "High".) The t-value (4.96) and the p-value (< 0.001) match what we calculated by hand above. There is also an R^2 value, which tells us how much of the variation in response variable (density) is explained by the predictor variable (elevation). In this case, approximately 55% of the variation is explained by elevation (This is rather high for one categorical value, maybe because I simulated these values).

### **You should now be able to complete problem 2 in the homework assignment**  
 
# Paired t-tests: Herbicide Treatments

## Data  

For this section of the lab, you will need to download the `caterpillar.csv` file from D2L. 

```{r, eval=FALSE}
caterpillar <- read.csv("data/caterpillar.csv")

```

Briefly, this is data from a controlled experiment looking herbicide effects on non-target caterpillars. Twelve plots were split, and each side was randomly assigned as a control or herbicide treatment. Each plot is an experimental unit and the treatments are "paired" within it.  

```{r}
head(caterpillar)
```

Plots were arranged in pairs at 12 different locations (`plot_number`). One plot in each pair was randomly selected for treatment with the microbial pesticide *Bacillus thuringiensis* (Bt) (`treatment`). The other plot was untreated and acts as a control. Surveys of non target caterpillars were performed by counting caterpillars on samples of 10,000 leaves on each plot (`caterpillar_count`). 

Because the plots were paired, what we're interested in is the difference in `caterpillar_count` between `bt_herbicide` and `control` plots. If the herbicide had no effect, then we would expect both plot types to have the same number of caterpillars. In other, words, the difference in counts between each plot would be 0. 

Therefore, the null hypothesis is that the average difference ($\mu_d$) between the two groups is 0. 


For a paired t-test, the null hypothesis is that the average difference between the two groups is 0. 

$$H_0:\large \mu_D = 0$$
And our alternative hypothesis is that there is a difference between the two groups, or:

$$H_A:\large \mu_D \ne 0$$

For this, we need to calculate the difference between the two treatments. The way that this data is organized, that means that we need to subtract row 2 from row 1, row 4 from row 3, etc. Based on the `dplyr` functions we've learned so far, there is no simple solution apparent.

Essentially, we calculate a one-sample t-test on the *differences* between groups. If the groups are the same, there should be no difference (i.e., difference = 0). If the groups are not the same, there should be a *difference*. 

We need to calculate the difference between the two treatments within each plot or experimental unit. If we look at the way the data is organized, there is not a simple solution using `dplyr` functions. 

```{r}
caterpillar
```

### Data Wrangling  

Luckily, there is a special helper function called `diff()` that we can use inside of `mutate()`. First, we need to group the data so that it uses the correct numbers. Here, we want to know what the effect is within plots, so we will use `group_by(plot_number)`. 

```{r}
caterpillar %>%
  group_by(plot_number) %>%
  mutate(difference = -diff(caterpillar_count))
```

We can see that this correctly calculated the difference between `treatment` types *within* `plot_number`, but it has doubled the number of observations (essentially giving a value for each plot:treatment combination, when we only want one observation per plot). We can modify our data pipeline above to `select()` the two columns we are interested in (`plot`, and `difference`) and then use `distinct()` to only keep unique, individual observations. In this step we will also save the output of this data pipeline as a new object which we can use for our formal analysis. 

```{r}
cat_diff <- caterpillar %>%
  group_by(plot_number) %>%
  mutate(difference = -diff(caterpillar_count)) %>%
  select(plot_number, difference) %>%
  distinct()
cat_diff
```

**REMEMBER** that when you are developing and working out a pipeline like this you should add only one command at a time. For example:

```{r, eval=FALSE}
# first step
caterpillar %>%
  group_by(plot_number)
# second step
caterpillar %>%
  group_by(plot_number) %>%
  mutate(difference = -diff(caterpillar_count))
# third step
caterpillar %>%
  group_by(plot_number) %>%
  mutate(difference = -diff(caterpillar_count)) %>%
  select(plot_number, difference)
# fourth step
caterpillar %>%
  group_by(plot_number) %>%
  mutate(difference = -diff(caterpillar_count)) %>%
  select(plot_number, difference) %>%
  distinct()

# Once you have the desired output, THEN save it as a new object
cat_diff <- caterpillar %>%
  group_by(plot_number) %>%
  mutate(difference = -diff(caterpillar_count)) %>%
  select(plot_number, difference) %>%
  distinct()
```

## Analysis  

Now that we have our a data object which includes the difference between treatments within each plot, we can formally test our hypotheses above using a linear model analysis. 

The structure of the t-test for paired data is:

$$\large t = \frac{\bar{y}_{diff} - 0}{SD_{diff} / \sqrt(n)}$$

For this section, we will skip the manual calculations and just use the `lm()` function. Also recall that the `lm()` function uses a more accurate "pooled" estimate of the standard deviation of the difference which is more tedious to calculate. Another good reason to use the `lm()` function - less chance to make mistakes!

But once again let's start by plotting our data. 

```{r}
ggplot(cat_diff, 
       aes(y = difference)) +
  geom_boxplot() +
  geom_hline(yintercept = 0,
             color = "dodgerblue",
             size = 1.5,
             linetype = "dashed") +
  theme_bw()
  
```

Because we already calculated the difference between the two treatments, we are fitting an "intercept-only" model as in the one sample t-test at the beginning of lab. However, in this case, our value of interest ($\mu_0$) is 0, which is the default in the `lm()` function, so we do not need to subtract anything from the response side of the formula. 

We can quickly plot these differences to get an idea of the overall effect. 

```{r}
cat_diff %>%
  ggplot(aes(y = difference)) +
  geom_boxplot(fill = "mediumorchid1") +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             size = 1.5) +
  theme_bw()
```

In this plot, the reference value ($\mu_0 = 0$) is represented with the dashed line, and is right on the edge of the upper quantile (75th percentile). It's difficult to tell if the differences are close enough to zero or not, so we will need to calculate a formal statistical test. 

Recall that we calculate a paired t-test statistic with the following formula:

$$\large t = \frac{\bar{y}_{Diff} - 0}{SED}$$

Here, SED is the standard error of the *difference*. For this example, we will skip the hand calculation and jump straight to using `lm()`. We already have the differences stored in the `cat_diff` object. We will use that object to calculate a one-sample t-test. Using the `lm()` function, a one sample t-test is the same as an intercept-only model.   

```{r}
paired_fit <- lm(difference ~ 1, data = cat_diff)
summary(paired_fit)
```


Our estimated intercept ($\pm$ SE) is 1.5 $\pm$ 0.65, and we can reject the null hypothesis and conclude that there was a difference in the number of caterpillars between the two treatments ($t_{statistic} = 2.32, df = 11, p = 0.04$). 

### **You should now be able to complete porblem 3 in the homework assignment.**  

## `lm()` analysis without data wrangling  

This section is supplementary and is not strictly required to complete HW 08.  

For completeness, and to match the example presented in Ch. 6 of *The New Statistics with R*, we could run a paired t-test on the original `caterpillar` data object. In this case, we would designate our response variable `caterpillar_count` and two predictor variables: `treatment` and `plot_number`. In this case, we are assessing the effect of treatment while controlling for the plot_number. This is also known as a "blocking" variable. We will return to analyses with multiple predictor or blocking variables when we discuss ANOVA and linear regression analyses.

Note that we modify the data to treat `plot_number` as a categorical variable using the `as.factor()` function. Becasue the variable is currently an integer, the `lm()` treats it as continuous which is not in fact correct. It's always important to think about the class or type of your individual variables to ensure you are running the analysis that you think you are running. 

```{r}
summary(lm(caterpillar_count ~ 
             treatment + as.factor(plot_number),
           data = caterpillar))
```

This output is much more crowded and difficult to interpret as it automatically runs a test comparing every `plot_number` to `plot_number == 1`. However, if we ignore those tests and focus on the second line (`treatmentcontrol`), we can see that it is the same results as our `paired_fit` object above. 


Our estimated intercept coefficient is -1.5, and the standard error of this estimate is 0.6455. The `summary()` output also includes a t~statistic~ (`t value`) of -2.324. The output does not give us our t~critical~ value, but it does gives us the p-value. Here, the p-value is 0.0403, which is less than our $\alpha$ value of 0.05. Hence, we reject the null hypothesis and conclude that there is a difference in treatment values. 

### **You should now be able to complete problem 3 in the homework assignment**  


## Coda

This concludes lab 08  
