---
title: "Graphing using ggplot"
subtitle: "ENVS475: Experimental Analysis and Design"
date: "`r format(Sys.time(), '%B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
acacia <- read.csv(here::here("data/uhuru_acacia.csv"))
```

## About

This lab will introduce you to the `ggplot2` package in R. This is a powerful and flexible package which is capable of producing publication quality grpahics. 

### Preliminaries

Before we begin, make sure to download the data from D2L. The data we are using in this lab is called `uhuru_acacia.csv`. You will also need to download `uhuru_trees.csv` for the homework assignment. Make sure to place both of these files in the `data/` folder in your R Project, otherwise you will need to modify the code below in order to have the code run successfully. 

### More information

> * You can view the [metadata](https://esapubs.org/archive/ecol/E095/064/metadata.php) from the *Acacia drepanolobium* Surveys. 

## Introduction to the UHURU dataset

### Data

* For the next set of lessons we'll be working with data on acacia size from an experiment in Kenya
* The experiment is designed to understand the influence of herbivores on vegetation by excluding different sized herbivores
* There are 3 different treatments:
  * The top-left image shows Megaherbivore  exclosures (MEGA), which use wires 2m high to keep out elephants
  * The top-right image shows Mesoherbivore exclosures (MESO), which use fenses starting 1/3m off the ground to exclude things like impalla
  * The bottom-left image shows full exclosures (TOTAL), which use fences all the way to the ground to keep out all mammalian herbivores
  * And the bottom-right image shows control plots (OPEN)

![Pictures of 4 experimental treatments. (A) Megaherbivore fences consist of two parallel wires starting 2-m above ground level. (B) Mesoherbivore fences consist of 11 parallel wires starting ~0.3 m above ground level and continuing to 2.4-m above ground level. (C) Total-exclusion fences consist of 14 wires up to 2.4-m above ground level, with a 1-m high chain-link barrier at ground level. (D) Open control plots are unfenced, with boundaries demarcated by short wooden posts at 10-m intervals.](https://esapubs.org/archive/ecol/E095/064/images/Fig2.png)

* So far we've been mostly working with data sets that are built into R, or making our own small data sets within our scripts. 
* Now we will need to read in the data sets into R. 
* But first, take a look at the `uhuru_acacia.csv` data sets by opening them. 
* * *Right-Click on uhuru_acacia.csv in your file explorer program to Open With Notepad or text editor*
* This data is comma separated, which is the default type in the `read.csv()` function.  
  * Recall that we used `read.csv()` in our `dplyr` homework. This is the main one we will throughout this course.  
  * You can also modify the `read.csv()` function to load other data types, such as tab-separated `.txt` files, etc.
  * This is beyond the scope of the course, but you can Google it to learn more  
* We'll call our data frame `acacia` and assign it the output from `read.csv()`
* We still to supply the file name in quotes as the first argument  
* to tell R that it's located in the `data/` folder, we add this to the beginning of the filename

```{r, eval=FALSE}
acacia <- read.csv("data/uhuru_acacia.csv")
```

* **Did everyone successfully load the data into R?**  
  + Are you in your R-project?  
  + Do you have a `data/` folder?  
  + Is your `uhuru_acacia.csv` in the `data/` folder?
  
  
Whenever you load a dataset into R, it is always a good idea to take a look at it to familiarize yourself with the structure. Run the following lines of code in RStudio one by one (I have removed the output to save space here): 
```{r, eval=FALSE}
names(acacia)
head(acacia)
tail(acacia)
str(acacia)
summary(acacia)
```

* You can see the entire `acacia` object by running the following command:  
`View(acacia)`  

* we can see that it includes information on:
  * the time and location of the sampling
  * the experimental treatment
  * the size of each Acacia including a height, the canopy diameter measured in the direction (or axis) of the largest diameter and the diameter of the axis perpendicular to that, and the circumference of the shrub  
  * information on the number of flowers, buds, and fruits
  * And finally information on the species of ant associated with the shrub. 
    + This is because there is a very interesting ant-acacia mutualism where the Acacia makes special structures that serve as houses for the ants and the ants swarm herbivores that try to eat the acacia  
  * **Note** that there are `NA` values in the HEIGHT column  
    * These refer to trees which were "dead"  

## `ggplot2`

* Very popular plotting package
* Good plots quickly
* Declarative - describe what you want not how to build it
* Contrasts w/Imperative - how to build it step by step
* Install `ggplot2` using `install.packages("ggplot2")`  
* Remember you only need to install packages on each machine once. 
  
### Basics

* We load the `ggplot2` package just like we loaded `dplyr` 

```{r, eval=FALSE}
library(ggplot2)
```

* If you haven't already loaded the uhuru data, make sure to do it now: 
```{r, eval=FALSE}
acacia <- read.csv("data/uhuru_acacia.csv")
```

#### `ggplot` basic syntax  

$$\underbrace{ggplot}_{initiate\; plot}(\underbrace{data = df}_{data\;frame},\; \underbrace{aes(x =\; , y = \;, color = \;)}_{plot\; attributes}) + \underbrace{geom\_line()}_{geometry}$$
* To build a plot using `ggplot` we start with the `ggplot()` function

```{r}
ggplot()
```

* `ggplot()` creates a base ggplot object that we can then add things to
  * Think of it as a blank canvas
  
* We can also add optional arguments for information to be shared across different components of the plot
* The two main arguments we typically use here are 
* `data` - which is the name of the data frame we are working with, so `acacia`
* `mapping` - which describes which columns of the data are used for different aspects of the plot
* We create a `mapping` by using the `aes` function, which stands for "aesthetic", and then linking columns to pieces of the plot
* We'll start with telling ggplot what value should be on the x and y axes
* Let's plot the relationship between the circumference of an acacia and its height

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT))
```
* Notice we now have "CIRC" on the x-axis and "HEIGHT" on the y-axis
* This still doesn't create a figure, it's just a blank canvas and some information on
  default values for data and mapping columns to pieces of the plot
* We can add data to the plot using layers
* We do this by adding a `+` after the the `ggplot()` function and then adding something called a `geom`, which stands for `geometry`
* To make a scatter plot we use `geom_point()`

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT)) +
  geom_point()
```

* It is standard to hit `Enter` after the plus so that each layer shows up on its own line
  
* To change things about the layer we can pass additional arguments to the `geom`
* We can do things like change
  * the `size` of the points, we'll set it to `3`
  * the `color` of the points, we'll set it to `"blue"`
  * the transparency of the points, which is called `alpha`, we'll set it to 0.5

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT)) +
  geom_point(size = 3, color = "blue", alpha = 0.5)
```

* To add labels (like documentation for your graphs!) we use the `labs` function

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT)) +
  geom_point(size = 3, color = "blue", alpha = 0.5) +
  labs(x = "Circumference [cm]",
       y = "Height [m]",
       title = "Acacia Survey at UHURU")
```


* Notice that when functions have lots of arguments, it is generally considered good style/formatting to hit `enter` after a comma. 
* For example, in the `labs()` call above, I hit enter after the `x = "Circumference [cm]",`
  * and after the `y = "Height [m]",`,
  * etc. 
* This makes your code **much easier** to read. 
  
### Rescaling axes

* We can rescale the axes to be on a `log10` scale using the 
  * `scale_y_log10()` and `scale_x_log10()` functions, respectively.
```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT)) +
  geom_point(size = 3, color = "blue", alpha = 0.5) +
  scale_y_log10() +
  scale_x_log10()
```

* **NOTE** This doe not change the data itself, just the presentation of it

> Do Problems 1.1 and 1.2 in homework 04: `ggplot`

### Grouping

#### Colors
* Visualize different groups on a single graph
* Look at influence of experimental treatment
* in `aes()`, we name a variable (column of the `acacia` data) to define the color in the plot

```{r}
ggplot(acacia,
       aes(x = CIRC,
           y = HEIGHT,
           color = TREATMENT)) +
  geom_point(size = 3, alpha = 0.5)
```

#### Facets

* Place all the data for one group in its own facet (i.e., panel)
* Facet specification

```{r}
ggplot(acacia, aes(x = CIRC, y = HEIGHT)) +
  geom_point(size = 3, alpha = 0.5) +
  facet_wrap(~TREATMENT)
```

* Where are all the acacia in the open plots? (eaten?)

> Do Problems 1.3 and 1.4 in homework 04: `ggplot`

### Statistical transformations

* We've seen that ggplot makes graphs by combining information on
  * Data
  * Mapping of parts of that data to aspects of the plot
  * A geometric object to represent the data

```{r}
ggplot(acacia, aes(x = CIRC, y = HEIGHT)) +
  geom_point()
```

* Many kinds of geometric object (type `geom_` and show completions)

* Each geom includes a statistical transformation
* So far we've only seen
    * `identity`: the raw form of the data or no transformation
* Transformations exist to make things like histograms, bar plots, etc.
* Occur as defaults in associated geoms

* To look at the number of acacia in each treatment use a bar plot
    * [`geom_bar()`](https://ggplot2.tidyverse.org/reference/geom_bar.html)

```{r}
ggplot(acacia, aes(x = TREATMENT)) +
  geom_bar()
```

* Uses the transformation `stat_count()`
    * Counts the number of rows for each treatment

* To look at the distribution of circumferences in the dataset use a histogram
    * [`geom_histogram()`](https://ggplot2.tidyverse.org/reference/geom_histogram.html)

```{r}
ggplot(acacia, aes(x = CIRC)) +
  geom_histogram(fill = "red")
```

* Uses `stat_bins()` for data transformation
    * Splits circumferences into bins and counts rows in each bin
* Set number of `bins` or `binwidth`
  * Default bin number is usually 30

```{r}
ggplot(acacia, aes(x = CIRC)) +
  geom_histogram(fill = "red", bins = 15)
```

```{r}
ggplot(acacia, aes(x = CIRC)) +
  geom_histogram(fill = "red", binwidth = 5)
```

* These can be combined with all of the other `ggplot2` features we've learned

> Do Tasks 2.1 and 2.2 in in homework 04: `ggplot`


#### Boxplots  

* [Boxplots](https://en.wikipedia.org/wiki/Box_plot) are a common way of summarizing data  
the `geom_boxplot()`

```{r}
ggplot(acacia, aes(x = TREATMENT,
                   y = HEIGHT)) +
  geom_boxplot()
```

### Position

* geom's also come with a default position
* In many cases the position is `"identity"`, which just means the object is plotted in the position determined by the data, but not always
* Let's remake our histogram, but color the bars by the treatment

```{r}
ggplot(acacia,
       aes(x = CIRC, color = TREATMENT)) +
  geom_histogram(binwidth = 5)
```

* You can see that the total height of the bars stayed the same
* That's because ggplot has just colored the pieces of each bar that correspond to each treatment
* It does this because the default `position` for `geom_histogram` is `"stacked"`, which stacks the bars on top of one another and therefore makes a stacked histogram.
* If we want separate overlapping histograms then we need to change the position

```{r}
ggplot(acacia, aes(x = CIRC, color = TREATMENT)) +
  geom_histogram(binwidth = 5, position = "identity")
```

* And then add some transparency so we can see all of the histograms

```{r}
ggplot(acacia, aes(x = CIRC, color = TREATMENT)) +
  geom_histogram(binwidth = 5, position = "identity", alpha = 0.5)
```


### Layers

* So far we've only plotted one layer or geom at a time, but we can combine multiple layers in a single plot
    * `ggplot()` sets defaults for all layers
    * Can combine multiple layers using `+`
    * The first geom is plotted first and then additional geoms are layered on top 

* Combine different kinds of layers
* Add a linear model

```{r}
ggplot(acacia, aes(x = CIRC, y = HEIGHT)) +
  geom_point() +
  geom_smooth(method = "lm")
```

* Both the `geom_point` layer and the `geom_smooth` layer use the defaults form
  `ggplot`
* Both use `acacia` for data and `x = CIRC, y = HEIGHT` for the aesthetic
* `geom_smooth` uses the statistical transformation `stat_smooth()` to produce a smoothed representation of the data
  * `method = "lm"` means a linear model.
  * The default is a non-linear method called "loess". 
    + definition of "loess" is beyond the scope of this course. 
  * Make the same plot as above again, this time without specifying the method. 
  
```{r}
ggplot(acacia, aes(x = CIRC, y = HEIGHT)) +
  geom_point() +
  geom_smooth()
```

* We also add linear models by a groups
  * Add linear model by treatment

```{r}
ggplot(acacia, aes(x = CIRC, y = HEIGHT, color = TREATMENT)) +
  geom_point() +
  geom_smooth(method = "lm")
```

* Because the color aesthetic is the default it is inherited by geom_smooth
* One set of points and one model for each treatment

### Multiple layers and posiiton

* summarize data and visualize individual observations  
  + Summary = `geom_boxplot()`
  + Individual points = `geom_point()`

```{r}
ggplot(acacia,
       aes(x = TREATMENT,
           y = HEIGHT)) +
  geom_boxplot() +
  geom_point()
```
* Some repeated observations in each TREATMENT group

* Add "noise" to observations with `position = position_jitter()`

```{r}
ggplot(acacia,
       aes(x = TREATMENT,
           y = HEIGHT)) +
  geom_boxplot() +
  geom_point(position = position_jitter())
```

* the default range of `position_jitter()` is usually too much.  
* Likewise, adding noise in both axes is misleading. 
  + in this case, only want to add noise to the horizontal, or x-axis, using the `width` argument  
  + remove noise from vertical (y-axis) by setting `height = 0`  
  
```{r}
ggplot(acacia,
       aes(x = TREATMENT,
           y = HEIGHT)) +
  geom_boxplot() +
  geom_point(
    position = position_jitter(
      height = 0,
      width = 0.1))
```

* add color by treatment  
* change the value of the `width` noise  
* make points transparent
* change the base "theme" of the plot  
  + I prefer `theme_bw()  
  + see [ggplot themes](https://ggplot2.tidyverse.org/reference/ggtheme.html) for a list of available themes  

```{r}
ggplot(acacia,
       aes(x = TREATMENT,
           y = HEIGHT,
           color = TREATMENT)) +
  geom_boxplot() +
  geom_point(
    alpha = 0.5,
    position = position_jitter(
      height = 0,
      width = 0.1)) +
  theme_bw()
```

### Changing values across layers

* We can also plot data from different columns or even data frames on the same graph
* To do this we need to better understand how layers and defaults work
* So far we've put all of the information on data and aesthetic mapping into `ggplot()`

```{r}
ggplot(
  data = acacia,
  # defaults set here for all subsequent layers
  mapping = aes(x = CIRC, 
                y = HEIGHT)) +
  geom_point() +
  geom_smooth(method = "lm")
```

* This sets the default data frame and aesthetic, which is then used by
  `geom_point()` and `geom_smooth()`
* Alternatively instead of setting the default we could just give these values
  directly to `geom_point()` and `geom_smo

```{r}
ggplot() +
  geom_point(
    data = acacia,
    mapping = aes(
      x = CIRC,
      y = HEIGHT,
      color = TREATMENT)) +
  geom_smooth(method = "lm")
```

* We can see that this information is no longer shared with other geoms since it is no longer the default, so we've asked for a linear model of nothing and so no model is shown

* Can use this to combine different aesthetics
* Make a single model across all treatments while still coloring points

```{r}
ggplot() +
  geom_point(
    data = acacia,
    mapping = aes(
      x = CIRC,
      y = HEIGHT,
      color = TREATMENT)) +
  geom_smooth(data = acacia,
              mapping = aes(
                x = CIRC,
                y = HEIGHT),
              method = "lm")
```

* `color` is only set in the aesthethic for the point layer
* So the smooth layer is made across all x and y values

* *Check if this makes sense to everyone*

* This same sort of change can be used to plot different columns on the same plot by changing the values of x or y
* If we wanted to plot two different columns as "Y" we could change the value of y in the aesthetic

* If we wanted to use data from two different data frames we could change the value of data

* We can also keep all of the values that are shared as defaults if we want to

```{r}
# set defaults in ggplot()
ggplot(data = acacia,
       mapping = aes(
         x = CIRC,
         y = HEIGHT)) +
  # add color mapping just for geom_point()
  geom_point(mapping = aes(
    color = TREATMENT)) +
  geom_smooth(method = "lm")
```

> Do Task 2.3-2.5 in in homework 04: `ggplot`

### Grammar of graphics

* Uniquely describe any plot based on a defined set of information
* Leland Wilkinson
* Geometric object(s)
  * Data
  * Mapping
  * Statistical transformation
  * Position (allows you to shift objects, e.g., spread out overlapping data points)
* Facets
* Coordinates (coordinate systems other than cartesian, also allows zooming)

### Saving plots as new files

* `ggsave()` function saves whatever the last plot you made was. 
* name the file you want to save:

```{r, eval=FALSE}
ggsave("acacia_by_treatment.jpg")
```
* The type of the file is determined by the file extension

```{r, eval=FALSE}
ggsave("acacia_by_treatment.pdf")
```

* if you want to save it in a specific folder, add it as `"folder_name\"` in front of the file name
  * Make sure you actually have a folder named that before saving!
* For example, if you followed the recommended file structure for this course, you should have a `homework` folder. 
* to save the last `ggplot` figure you made as a .png file to your homework folder, enter the following:
```{r, eval = FALSE}
ggsave("homework/acacia_by_treatment.png")
```

* Check your folder to make sure the file is there. 
* Open it to see how it looks
* You may want to delete this from your homework folder so you don't get confused later. 

#### Other `ggsave()` arguments

Enter `?ggsave()` in the console to see the help file for this function. This lists the arguments that you can enter and gives a brief explanation of them

* Size
* define the height and width of your image

```{r, eval=FALSE}
ggsave("acacia_by_treatment.pdf", height = 5, width = 5)
```

* define the measurement unit
```{r, eval=FALSE}
ggsave("acacia_by_treatment.pdf",
       height = 5,
       width = 5,
       units = "cm")
```

* set it to inches:
```{r, eval=FALSE}
ggsave("acacia_by_treatment.pdf",
       height = 5,
       width = 5,
       units = "in")
```

* The default is to save the `last_plot()` that you created
  * See help file
* if you save your ggplot as an object, you can tell `ggsave()` which object you want to save. 

```{r, eval=FALSE}
# histogram
acacia_hist <- ggplot(acacia,
                      aes(x = CIRC)) +
  geom_histogram(fill = "red", binwidth = 5)

# scatter plot
acacia_point <- ggplot(acacia,
                      aes(x = CIRC,
                          y = HEIGHT)) +
  geom_point()

# linear model
acacia_lm <- ggplot(
  data = acacia,
  mapping = aes(x = CIRC, 
                y = HEIGHT)) +
  geom_point() +
  geom_smooth(method = "lm")

# save the histogram plot

ggsave("acacia_histogram.png",
       plot = acacia_hist)
```
  
> Complete HW 04: `ggplot`