---
title: "LAB: Graphing using ggplot"
subtitle: "ENVS475: Experimental Analysis and Design"
date: "`r format(Sys.time(), '%B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
invert <- read_csv(here::here("data/ggplot_lab_invert_survey.csv"))
```

## About

This lab will introduce you to the `ggplot2` package in R. This is a powerful and flexible package which is capable of producing publication quality graphics. 

### Preliminaries

Before we begin, make sure to download the data from D2L. The data we are using in this lab is a small simulated data set called `ggplot_lab_invert_survey.csv`. You will also need to download `galapago-finches.csv` for the homework assignment. Make sure to place both of these files in the `data/` folder in your R Project, otherwise you will need to modify the code below to have it run correctly. 

* `ggplot2` is a part of the `tidyverse` and was included when you previously ran `install.packages("tidyverse")`  

* Remember you only need to install packages on each machine once.  

```{r, eval=FALSE}
library(tidyverse)
```

## Getting Started  

* So far we've been mostly working with data sets that are built into R, or making our own small data sets within our scripts. 
* Now we will need to read in the data sets into R. 
* But first, take a look at the `ggplot_lab_invert_survey.csv` data set by opening them. 
* * *Right-Click on invert.csv in your file explorer program to Open With Notepad or text editor*  
* This data is comma separated, which is the default type in the `read_csv()` function. `read_csv()` is a part of the `tidyverse` and is the the main function that we will throughout this course.  
  * There is also a `read.csv()` function that comes with the base installation of R. It behaves very similarly but is slightly less "smart" than `read_csv()`. 
* We'll call our data frame `invert` and assign it the output from `read-csv()`
* We still to supply the file name in quotes as the first argument  
* to tell R that it's located in the `data/` folder, we add this to the beginning of the filename

```{r, eval=FALSE}
invert <- read_csv("data/ggplot_lab_invert_survey.csv")
```

* **Did everyone successfully load the data into R?**  
  + Are you in your R-project?  
  + Do you have a `data/` folder?  
  + Is your `ggplot_lab_invert_survey.csv` in the `data/` folder?
  
  
Whenever you load a dataset into R, it is always a good idea to take a look at it to familiarize yourself with the structure. Run the following lines of code in RStudio one by one (I have removed the output to save space here): 
```{r, eval=FALSE}
names(invert)
head(invert)
tail(invert)
str(invert)
summary(invert)
```

* You can see the entire `invert` object by running the following command:  
`View(invert)`  

* we can see that it includes information on:
  * the month (as a number) of the sampling month  
  * the taxa (either `beetle` or `spider`) as a character vector 
  * the body length (in millimeters) of each individual   
  * and the body mass (in grams) of each individual  
  
# `ggplot2`

* Very popular plotting package  

* Makes publication quality plots quickly  

* Declarative - describe what you want not how to build it  

* Contrasts w/Imperative - how to build it step by step  
  
## Basics

### `ggplot` basic syntax  

$$\underbrace{ggplot}_{initiate\; plot}(\underbrace{data = df}_{data\;frame},\; \underbrace{aes(x =\; , y = \;, color = \;)}_{plot\; attributes}) + \underbrace{geom\_line()}_{geometry}$$
* To build a plot using `ggplot` we start with the `ggplot()` function

```{r}
ggplot()
```

* `ggplot()` creates a base ggplot object that we can then add things to  

  * Think of it as a blank canvas
  
### Plot attributes, mapping and `aes()`
  
* We can also add optional arguments for information to be shared across different components of the plot
* The two main arguments we typically use here are 
* `data` - which is the name of the data frame we are working with, so `invert`
* `mapping` - which describes which columns of the data are used for different aspects of the plot
* We create a `mapping` by using the `aes` function, which stands for "aesthetic", and then linking columns to pieces of the plot
* We'll start with telling ggplot what value should be on the x and y axes
* Let's plot the relationship between the `body_length_mm` of invertebrates to their `body_mass_g`  
* note that I start a new line after each `,` and that Rstudio automatically indents for us. This improves readability and makes it easier to look for typos.  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g))
```

* Notice we now have "CIRC" on the x-axis and "HEIGHT" on the y-axis
* This still doesn't create a figure, it's just a blank canvas and some information on the default values for the data and mapping columns to axes of the plots  

### Plot Layers  

* We can add data to the plot using layers  

* We do this by adding a `+` after the the `ggplot()` function and then adding something called a `geom`, which stands for `geometry`  

* To make a scatter plot we use `geom_point()`

```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g)) +
  geom_point()
```

* It is standard to hit `Enter` after the plus so that each layer shows up on its own line
  
* To change things about the layer we can pass additional arguments to the `geom`  

* We can do things like change  

  * the `size` of the points, we'll set it to `3`  
  
  * the `color` of the points, we'll set it to `"blue"`  
  
  * the transparency of the points, which is called `alpha`, we'll set it to 0.5  
  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g)) +
  geom_point(size = 3, color = "blue", alpha = 0.5)
```

* To add labels (like documentation for your graphs!) we use the `labs` function  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g)) +
  geom_point(size = 3, color = "blue", alpha = 0.5) +
  labs(x = "Body length [mm]",
       y = "Body mass [g]",
       title = "Simulated ground-dwelling invertebrate survey")
```

# One Continuous Variable  

Throughout the semester, we will want to see the distribution of one continuous variable. The two main ways we will do that is through:  

1. Histograms  

2. Boxplots 


## Histograms  

* We can make a histogram of `body_mass_g` by modifying our earlier code.  

* Note only **one** variable, so we remove the `y=...` part.  

* We also change `geom_point()` to `geom_histogram`  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g)) +
  geom_histogram()
```

* `geom_histogram()` groups observations into bins  

  * i.e., all the numbers between 10-12 g  
  
* automatically picks bin numbers based on data (graph above used 30 bins, see message output), but we can modify this using the `bins` argument   


### 5 bins  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g)) +
  geom_histogram(bins = 5)
```

* see that there are now 5 bins, but the "edges" of the bins are unclear    

### Binwidth  
* We can set the size of the bins with `binwidth` argument   

* for example, let's say we want bins that are 2.5 g across  


```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g)) +
  geom_histogram(binwidth = 2.5)
```

### Histogram attributes  

* We can change the attributes of the histogram output in similar ways as above  

  * Common arguments are: `fill`, `color` (outline), `size` (outline), and `alpha` (transparency).  
  
  * You can also explore different "themes". Below is `theme_bw()`, but you can try `theme_classic()`, `theme_dark()`, `theme_void()`, etc.   
  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g)) +
  geom_histogram(binwidth = 2.5,
                 fill = "dodgerblue",
                 color = "black",
                 size = 2,
                 alpha = 0.5) +
  theme_bw()
```


## Boxplots  

* Boxplots summarize the data distribution  

* AKA "Cat and Whisker" Plots  


```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g)) +
  geom_boxplot()
```
* whiskers: show minimum (left side) and maximum (right side) of data  

* Box: shows where 50% of the data lies  

* Line in middle of box is the median  

* Useful to loo for normal distributions:  

  * Symmetry of box, whiskers, etc.   
  

### Boxplot attributes  
```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g)) +
  geom_boxplot(fill = "red",
               color = "black")
```

## Density Plots 

* Density plots show the full distribution  

* i.e., a histogram has discrete bins, but a density plot is continuous  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g)) +
  geom_density(fill = "pink",
               alpha = 0.75) +
  theme_dark()
```


**You should now be able to complete problem 01 in the homework assignment**

# One Continuous Variable, and Categories  

## Boxplots  

We will often want to compare the distributions between two groups. We can do this by modifying our boxplot code above by adding a grouping variable as our `y = ...` in the `aes()` function  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g,
                     y = taxa)) +
  geom_boxplot()
```

* we can also vary the color/fill by taxa by including it in the `aes()` function  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g,
                     y = taxa,
                     fill = taxa)) +
  geom_boxplot()
```

* ggplot automatically picked two colors for us based on the values in the `taxa` column  


### Change fill colors  

### Manually  

See a [list of colors](https://derekogle.com/NCGraphing/resources/colors) available in ggplot  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g,
                     y = taxa,
                     fill = taxa)) +
  geom_boxplot() +
  scale_fill_manual(values = c("dodgerblue", "hotpink"))
```

* note that this **does not** work if you put it in the `geom_boxplot()` argument:  

```{r, eval=FALSE}
ggplot(data = invert,
       mapping = aes(x = body_mass_g,
                     y = taxa)) +
  geom_boxplot(fill = taxa)
```

* Likewise, if you put a fill/color argument in the `geom_boxplot()` function, it will overwrite what you mapped above  


```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g,
                     y = taxa,
                     fill = taxa)) +
  geom_boxplot(fill = "black")
```

### Categorical vs. continuous variables  

* Our data has the results of three surveys across months   

* we may want to see if there are different distributions across months   


```{r}

ggplot(data = invert,
       mapping = aes(x = body_mass_g,
                     fill = month)) +
  geom_boxplot()
```

* this didn't work because `month` is a number, which R interprets as a continuous variable.  

* notice the warning discusses "group" structure of the data  

* we can get around this a number of ways.  

* the easiest at this point is to force month to be a `factor` (what R calls a categorical variable).  


```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g,
                     fill = as.factor(month))) +
  geom_boxplot()
```


## Histograms 

* We can also do a similar thing with histograms.  

* note that we need to **remove** the `y=...` argument, and we also need to add `position = "identity"` to the histogram call.  

  * otherwise the counts are "stacked" and the display is misleading

```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g,
                     fill = taxa)) +
  geom_histogram(position = "identity",
                 alpha = 0.75) +
  scale_fill_manual(values = c("grey30", "goldenrod")) +
  theme_bw()
```


**You should now be able to complete problem 02 in the homework assignment**


# Two Continuous Variables  

## Scatter Plots  
* Display two continuous variables by mapping them to `x` and `y`  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g)) +
  geom_point(size = 3)
```

### map attributes to other columns  

#### Color  

* Using a categorical variable  

* Gives discrete colors  

  * i.e., "red", or "blue"  
  
  
```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g,
                     color = taxa)) +
  geom_point(size = 3)
```

* Using a continuous variable  

* gives "shades" of one color, based on value  


```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g,
                     color = month)) +
  geom_point(size = 3)
```

#### Shape  
```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g,
                     shape = taxa)) +
  geom_point(size = 4)
```

#### Shape + color  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g,
                     color = taxa, 
                     shape = as.factor(month))) +
  geom_point(size = 4)
```


## `geom_smooth()`

*  We will often want to estimate the relationship between our predictor (`x`) and our response (`y`).  

* do this with `geom_smooth()`

```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g)) +
  geom_point(size = 4) +
  geom_smooth()
```

* the default is a "squiggly line"  

* we will limit ourselves to linear relationships, so add `method = "lm"`  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm")
```

* `geom_smooth()` inherits other aspects from aes.  

* add a smooth for both taxa:  


```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g,
                     color = taxa)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm")
```

## Facets  

In any of the plots, it is also possible to add facets, which gives each group their own panel:  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_length_mm,
                     y = body_mass_g,
                     color = taxa)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm") +
  facet_wrap(~taxa)
```


```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g,
                     y = taxa,
                     fill = taxa)) +
  geom_boxplot() +
  scale_fill_manual(values = c("dodgerblue", "hotpink")) +
  facet_wrap(~taxa)
```

* can also add a different variable  

* this helps to visualize when we have 2+ grouping variables  


```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g,
                     y = taxa,
                     fill = taxa)) +
  geom_boxplot() +
  scale_fill_manual(values = c("dodgerblue", "hotpink")) +
  facet_wrap(~month)
```
> Do Tasks XXX in the homework for this week 

## Coda  

**You should now be able to finish the homework. The below information is provided for your personal edification** 

### Grammar of graphics

* Uniquely describe any plot based on a defined set of information  

* Leland Wilkinson  

* Geometric object(s)  

  * Data  
  
  * Mapping  
  
  * Statistical transformation  
  
  * Position (allows you to shift objects, e.g., spread out overlapping data points)  
  
* Facets  

* Coordinates (coordinate systems other than cartesian, also allows zooming)  


### Saving plots as new files

* `ggsave()` function saves whatever the last plot you made was.  

* let's rerun this code:  

```{r}
ggplot(data = invert,
       mapping = aes(x = body_mass_g,
                     y = taxa,
                     fill = taxa)) +
  geom_boxplot() +
  scale_fill_manual(values = c("dodgerblue", "hotpink")) +
  facet_wrap(~month)
```

* we will now save it into the home directory of our project and name the file:

```{r, eval=FALSE}
ggsave("invert_by_month_taxa.jpg")
```

* The type of the file is determined by the file extension  

```{r, eval=FALSE}
ggsave("invert_by_month_taxa.jpg")
```

* if you want to save it in a specific folder, add it as `"folder_name\"` in front of the file name  

  * Make sure you actually have a folder named that before saving!  
  
* For example, if you followed the recommended file structure for this course, you should have a `homework` folder.   

* to save the last `ggplot` figure you made as a .png file to your homework folder, enter the following:  

```{r, eval = FALSE}
ggsave("homework/invert_by_month_taxa.png")
```

* Check your folder to make sure the file is there.  

* Open it to see how it looks  

* You may want to delete this from your homework folder so you don't get confused later. 
