---
title: "Exploring data with `dplyr` "
subtitle: "ENVS475: Experimental Analysis and Design"
date: "`r format(Sys.time(), '%B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)

library(dplyr)
```

# Overview 

This tutorial is aimed at giving you an introduction to the popular `dplyr` package. `dplyr` provides a few simple verbs that allow you to quickly and easily select and manipulate your data, and create an interactive environment for data exploration. This tutorial is based on Chapter 5 in ["*R for Data Science*"](http://r4ds.had.co.nz) by Hadley Wickham & Garrett Grolemund which is vailable for free online. Material was also adapted from the dplyr introduction vignette which you can find by running the following command:
```{r eval = FALSE}
vignette("dplyr")
```

`dplyr` is part of the suite of packages known as `tidyverse`. The packages within the tidyverse were designed with a similar data and programming philosophy, and work together fluidly. Hadley Wickham is the lead developer of the tidyverse, and his book above takes a tidyverse-centric view, and is a fantastic resource if you want to learn more. We will be using the `tidyverse` packages throughout this course. 

## Preliminaries

Before we begin, make sure you have the `tidyverse` installed. 

If you need to install it, run the following lines of code, as needed. Also note that the package name is inside of "" in the function call: 
```{r, eval=FALSE}
install.packages("tidyverse")
```

As a reminder, you only have to run the `install.packages(...)` line once per machine. Running this line downloads the package onto the machine that you are currently using. However, you do need to load the packages at the beginning of every session. The following commands "loads" the packages and makes the commands available to you. Note that once the package is installed, you *do not* need "" around the package name.

```{r eval = FALSE}
library(tidyverse)
```

## Data: `starwars`  

In order to demonstrate the basic data manipulation verbs of dplyr, we will be using the `starwars` database, which comes with `dplyr` (part of the tidyverse. 

We will load the `starwars` data by running the following command:
```{r}
data("starwars")
```

Whenever you start working with a new dataset, you should also run some quick commands to get an idea of what the data looks like, and how it is set up. I generally run the following commands (note that I have suppressed the display in this document for room, but you should run each of these on your computer):

```{r, eval=FALSE}
dim(starwars)
names(starwars)
head(starwars)
tail(starwars)
str(starwars)
```

The `dim()` list the number of rows and columns, in that order. The `starwars` dataset has `r nrow(starwars)` rows and `r nrow(starwars)` columns. 

You may notice that the `head(starwars)` output is different from many other data frames that you may have worked with before. It only shows the rows and columns that easily fit into your window. That's because the `starwars` object is a tibble:

```{r}
class(starwars)
```

Tibbles `"tbl_df"` act just as a data frame would, but they print only the information that can fit into your window. To see the full data frame in the Rstudio viewer use:

```{r eval = FALSE}
View(starwars)
```

Or as a normal data frame use:

```{r eval = FALSE}
head(as.data.frame(starwars))
```

To convert a data frame to a tibble, use `as_tibble()`. 

As a general rule, I almost always use the `names()` function when I'm looking at a new data set. This helps me get an idea of the data structure, and if the variable names are meaningful, can help me understand what types of questions I can ask, and how I may be able to answer them.

```{r}
names(starwars)
```

These variables should be relatively self-explanatory, but you can run `?starwars` for more information, including units. 


#  `dplyr`

Data manipulation is one of the primary tasks that scientists undertake. the `dplyr` package makes it easy to select the data you want, organize it in a useful way, calculate useful new variables, and summarize data by groups. 

There is an extremely helpful Rstudio data wrangling cheat sheet which you can download directly through Rstudio by clicking *Help > Cheatsheets > Data transformation with dplyr*. 

**Practice good project organization** When you download these cheat sheets, put them all in an appropriate folder so you can find them again easily. Recall that I recommend the following sub folders in your `ENVS_475` project directory:
```
ENVS_475/
  |─── other_resources/
        |─── cheatsheets/
              |─── data-transformation.pdf
```

I almost always have this cheat sheet open while I'm conducting any analyses. It takes a little effort to learn how to "read" the cheat sheet, but it's well worth the effort. 

 
## Single table verbs

dplyr aims to provide a function for each basic verb of data manipulation:

* Pick observations based on their values: `filter()`  
  * many rows to fewer rows  
* Reorder the rows: `arrange()`  

* Pick variables based on their names: `select()`  
  * many columns to fewer columns  
  
* Add new variables (columns) based on existing variables: `mutate()` 
  
* Reduce many observations to a single summary: `summarise()`  
  * many rows to one  

* Combine `summarise()` with `group_by()` to get one row per group  

All verbs work similarly and have similar syntaxes:

1. The first argument is a data frame.
2. Following arguments specify what to do with the data frame using un-quoted variable names. 
3. Result is new data frame. 

There are other verbs in the `dplyr` package, along with many useful "helper" functions. Browse the vignettes for dplyr: `browseVignettes(package = "dplyr")`, or check the Rstudio data wrangling cheat sheet which you downloaded at the beginning of this tutorial. 

### Filter rows with `filter()`

##### Many rows to fewer rows  

`filter()` allows you to select a subset of rows in a data frame. The first argument is the name of the data frame. The second and subsequent arguments are the expressions that filter the data frame. 

#### Filter rows based on numeric data  

For example, we can select all star wars characters who are greater than or equal to 180 cm tall with:
```{r}
filter(starwars, height >= 180)
```

* Note that R tells us that the functioned returned "A tibble: rows x columns".  

*  This makes sense, because we are only asking for star wars characters taller than a number.   

* It's always a good idea to double check that the size of the returned data makes sense. 


#### The use of `==`  

if we wanted to know which characters were *exactly* 180 cm, we use the following: 

```{r}
filter(starwars, height == 180)
```

In the `filter()` command, we use `==` instead of `=`. The `==` is a Boolean operator. It is used to specify that we want to know if the object on the left is *exactly* the same as the object on the right. For example, run the following code:

```{r}
5 == 5
5 == 9
"A"=="a"
"a" == "a"
```

The single `=` is another way of *assigning* a value to an object in R. For example `a = 5` stores the value of 5 inside of the object called `a`. 

#### Filtering rows based on character data

In the example above, we were filtering observations based on numeric data (i.e., `height >= 150`). Filter also works with character data types. For example, let's extract all of the star wars characters that are from Naboo, which is stored in the `homeworld` column

```{r}
filter(starwars, homeworld == "Naboo")
```
Note the use of `""` around UA in the `filter()` function. Remove the quotes and run the code again. What happens?

You should have received an error code in the R console. Error codes often look like hieroglyphics the first few times you see them, but the important information in this error is the "object 'UA' not found". When we enter unquoted arguments into the dplyr functions, R looks for objects (i.e., data objects, column names within a data object, etc.) with that name. 


#### Other Boolean operators  

Other Boolean operators include:
* `|` which means "or"  

* `&` which means "and"  

* `!=` which means "not equal to"  **PAY ATTENTION HERE!**

* `>=`which means "greater than or equal to"  

* `<=`which means "less than or equal to"  

* `>`which means "greater than"  

* `<`which means "less than"  


You can use boolean operators, singly and in combination:

```{r, eval = FALSE}
# starwars from Tatooine *OR* Alderaan
filter(starwars, homeworld == "Tatooine" | homeworld == "Alderaan")

# starwars from Tatooine OR Alderaan AND that are > 100 OR < 150 cm
filter(starwars, homeworld == "Tatooine" | homeworld == "Alderaan" & height > 100 | height < 150)
```

Sometimes you want to see all of the data that do *not* meet a criteria. For example, maybe you don't want to know what starwars characters are not from Tatooine. You can use the logical negation command `!=`. This command can be read as "not equal to". 

```{r}
# Not Tatooine
filter(starwars, homeworld != "Tatooine")
```

#### Removing `NA`'s with `filter()`

A common task in data analysis requires the removal of `NA` values in our data. For example, in the `starwars` data, droids don't have hair, so `hair_color` is `NA`. 

Based on what we learned before it’s natural to think that we do this by using the condition `hair_color != NA`
* To remove only rows with `NA` in specific columns use `filter`

```{r, eval=FALSE}
filter(starwars, hair_color != NA)
```

* Why didn't that work?
* Null values like `NA` are special
* We don't want to accidentally say that two "missing" things are the same
    * We don't know if they are
* So use special commands
* `is.na()` checks if the value is `NA`
* Combine this with `!` for "not"

```{r}
filter(starwars, !is.na(hair_color))
```

#### Using `unique()` 

It is also often useful to see all of the values that occur within a variable column. For example, what are all of the homeworld origins in this data set?

Here, I will use the base R function `unique()` to see all of the values in the variable `homeworld`. To select just that variable, I will use the `$` subsetting function. 
```{r}
unique(starwars$homeworld)
```

### Arrange rows with `arrange()`  

##### Change order of rows based on values in a column  

`arrange()` has a similar syntax as `filter()` but instead of selecting rows, it reorders them. It takes a data frame as the first argument, and then column name(s) to order by as subsequent arguments. The data will be arranged by the first column name provided, with ties being broken by subsequent columns:


```{r}
# arrange by homeworld
arrange(starwars, homeworld)

# arrange by homeworld, then height
arrange(starwars, homeworld, height)
```

The default order is from smallest to largest, numeric to character, and a to z. To reverse this order, use `desc()` to order a column in descending order:

```{r}
arrange(starwars, desc(homeworld))
```

### Select columns with `select()`  

##### Many columns to fewer columns

Data sets generally contain numerous columns, but oftentimes you are only interested in a few for a given analysis. `select()` allows you to focus on a useful subset of your data while dropping un-needed columns. You can indicate which columns you want by naming them (without quotes) or select which columns you *don't* want by putting a `-` in front of the name. I have suppressed the output to save space in this document, but be sure to run the following commands in your session and see what happens. Be sure to note the number of rows and columns that each command returns.

```{r, eval=FALSE}
# Select columns by name
select(starwars, height, mass, homeworld, hair_color)

# Select all columns between name and eye_color (inclusive)
select(starwars, name:eyecolor)

# select all columns except for height
select(starwars, -height)
```


There are a number of helper functions you can use within `select()`, like `starts_with()`, `ends_with()`, `matches()` and `contains()`. These let you quickly match larger blocks of variables that meet some criterion. See `?select` for more details.


The order you write the arguments in `select()` are retained in the output. This can be helpful if you want to quickly compare the values in two columns which do not normally appear next to each other. 

``` {r}
select(starwars, name, homeworld, eye_color)
```

### Add new columns with `mutate()`  

##### Many columns to more columns  

Data analysis often requires the creation of new variable columns based on values within your data. The `mutate()` function allows you to do this. For example, let's calculate a new variable called `bmi`, which looks at how height and mass in one variable.

```{r}
# make new column called "bmi"
d2 <- mutate(starwars,
  bmi = mass / height^2)

# select new columns in d2, and columns used to calculate them
select(d2, mass, height, bmi)
```


### Summarise values with `summarise()`  

##### Many rows to a single row  

The last verb is `summarise()`. This collapses a dataframe to a single value, based on a function:

```{r}
# mean height
summarise(starwars,
  height_average = mean(height, na.rm = TRUE))
# smallest mass
summarise(starwars, min_mass = min(mass, na.rm = TRUE))
```

Notice the argument `na.rm = TRUE` within the summarize function. If you have missing or `NA` values within your data, it will cause the summary functions to return `NA`. Recall that we discussed this in the "`NULL`'s in Vectors " section in our introduction to R activity.  

Additionally, summarize is optimized to work with functions that return a single value. For example, `range()` returns the minimum and maximum value of a set of numbers:
``` {r}
range(c(1, 2, 3, NA, 5, 6, 7), na.rm = TRUE)
```

When used in combination with `summarise()`, two values are returned, but they are not labeled, and it includes a warning message:
```{r}
summarise(starwars,
          height_range = range(height, na.rm = TRUE))

```

If this is the only thing you are calling, it may be easy to tell which is which, but if we have multiple summary arguments, it can be less obvious:

```{r}
summarise(starwars,
          height_range = range(height, na.rm = TRUE),
          height_mean = mean(height, na.rm = TRUE))

```

However, we can get around this by calling the `min()` and `max()`  functions separately within summarise:
``` {r}
summarise(starwars, 
          min_height = min(height, na.rm = TRUE),
          max_height = max(height, na.rm = TRUE),
          mean_height = mean(height, na.rm = TRUE))
```


## Grouped operations  

##### Many rows to one-per-group  

These verbs are useful on their own, but when used in conjuction with the `group_by()` function, the awesomeness of `dplyr` starts to shine through. It organizes a dataset into specified groups of rows. Verbs are then applied group-by-group within the dataset. Conveniently, this is accomplished using the exact same syntax as above. 

Here, we will group the data by `homeworld` and then summarize the mean height:
```{r}
# make a new object which contains a "grouping variable"
home_group <- group_by(starwars, homeworld)
summarize(home_group,
          mean_height = mean(height, na.rm = TRUE))
```
The output now contains 1 row per world 

Notice that if we print out the `home_group` object that it has the same dimensions as the original, but there is now a new line at the top which specifies the grouping variable:

```{r}
home_group
```

#### Using `n()` to get counts

Another common data summarization task is to get the number of observations within a data set. For example, let's say we wanted to know how many starwars each carrier was scheduled to operate. For this, we can use `n()`, which is a special function inside of the `summary()` function. 

```{r}
summarize(home_group, count = n())
```


# Chaining

The `dplyr` function calls don't have any side-effects (unlike some base functions), making it easy to explore your data in an interactive way. However, one disadvantage of this is it doesn't lead to very sucinct code, particularly if you want to perform many operations at once. You can do it step-by-step, saving a new object each time:

```{r, warning = FALSE, message = FALSE, eval = FALSE}
# calculate starwars characters that have a BMI either >0.04 or < 0.01 
a1 <- group_by(starwars, homeworld)
a2 <- select(a1, homeworld, height, mass)
a3 <- mutate(a2,
  bmi = mass / height^2)
a4 <- filter(a3, bmi > 0.04 | bmi < 0.005)
a4
```

However this can lead to many problems. Giving objects appropriate names can be difficult (e.g. object names in `ggplot` example). When naming them chronologically (as above), it can be difficult to remember which object is which (was the summarized object `a2` or `a3`?...). Especially if you want to do the same thing over and over on different data sets or subsets of observations. 

## The pipe operator: `%>%`, `|>`

The pipe operator is originally from the `magrittr` package, but is included automatically in the `dplyr` package. More recently, newer version of R include a "native pipe" `|>` as a part of the base install. Both pipes should work the same, and either is fine for this class. I learned using the `%>%` pipe, so most of my code will use this one. 

* The pipe takes the output of one command and passes it as an input to the next command. 

* The pipe operator allows you to write this function: `f(x,y)` as `x %>% f(y)`. 

I think that this is a bit confusing to think about, but when you see some examples the power and ease of use becomes obvious. 

```{r}
# a silly example
x <- seq(10)
max(x)
# is the same as
x %>% max()

mean(x)
x %>% mean()
```

You can see that both syntaxes give the same result. Now, let's go back our example looking at arrival and departure delays > 30 minutes

```{r, message = FALSE, eval = FALSE}
# original method, saving new object at each step
a1 <- group_by(starwars, homeworld)
a2 <- select(a1, homeworld, height, mass)
a3 <- mutate(a2,
  bmi = mass / height^2)
a4 <- filter(a3, bmi > 0.04 | bmi < 0.005)
a4

# now using the %>% operator
a5 <- starwars %>%
  group_by(homeworld) |>
  select(homeworld, height, mass) |>
  mutate(bmi = mass / height^2) |>
  filter(bmi > 0.04 | bmi < 0.005)

identical(a4, a5)
#> TRUE
```

It is helpful to say "then" when you see a `%>%` operator. 
e.g. take object `starwars`, *then* group by year, month and day, *then* select variables arr_delay and dep_delay *then* summarize ... etc.

The pipe operator in combination with `dplyr` allows you to quickly examine your data and explore interesting results. One of my favorite aspects of this is you can answer interesting questions almost immediately. 

"I wonder what the smallest mass character is?"

```{r}
starwars %>% 
  arrange(mass)
```

If you want to also know the homeworld:
```{r}
starwars %>% 
  arrange(mass) %>%
  select(name, mass, homeworld)
```

What homeworld has the largest average mass?

```{r, message = FALSE}
starwars %>% 
  group_by(homeworld) %>%
  summarise(mean_mass = mean(mass)) %>%
  arrange(desc(mean_mass))
```

## Formatting with pipes

A couple of pointers.

  * always enter a new line after a pipe `%>%`
  
  * if you have many arguments within a function in a pipe, enter a new line after each comma
  
  * if you are making a new variable with `mutate`, or summarizing variables with `summarize`, give them a meaningful name
  
  * try and limit the number of pipes in a single call. 5-6 is OK, but 10 or more should be avoided. If you need that many, save an intermediate object and then pipe that. 

#### Bad formatting
```{r, echo=TRUE, eval = FALSE, warning=FALSE, message=FALSE}
# don't do these:
# no new line after each pipe
starwars %>% group_by(year, month, day) %>% filter(carrier == "FL" | 
carrier == "AA"| carrier == "UA") %>% select(arr_delay, dep_delay) %>%
summarise(arr = mean(arr_delay, na.rm = TRUE), dep = mean(dep_delay, na.rm
= TRUE)) %>% filter(arr > 30 | dep > 30)
# no new line after each summarize argument, and meaningless variable names
starwars %>% summarize(x1 = mean(dep_delay, na.rm = TRUE), x2 = min(dep_delay, na.rm = TRUE), x3 = max(dep_delay, na.rm = TRUE), x4 = n())
```

#### Better formatting
```{r, echo=TRUE, eval = FALSE, warning=FALSE, message=FALSE}
# new line after each pipe
starwars %>%
  group_by(year, month, day) %>%
  select(arr_delay, dep_delay) %>%
  summarise(arr = mean(arr_delay, na.rm = TRUE), # new line here
            dep = mean(dep_delay, na.rm = TRUE)) %>%
  filter(arr > 30 | dep > 30)

# new line after each summarize argument, and variable names
starwars %>%
  summarize(mean_delay = mean(dep_delay, na.rm = TRUE),
            min_delay = min(dep_delay, na.rm = TRUE),
            max_delay = max(dep_delay, na.rm = TRUE),
            n_samples = n())
```

Note that when you press enter when "inside" of a function, RStudio automatically adds a tab space. This is very helpful for keeping the code format clean and makes it much easier to read. 

