---
title: "Homework 3: dplyr"
subtitle: "ENVS475: Experimental Analysis and Design"
date: "`r format(Sys.time(), '%B, %Y')`"
output: 
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
   warning = FALSE, message = FALSE,
   eval = FALSE,
   fig.align = 'center', fig.retina = 2,
  collapse = TRUE,
  comment = "#>"
)
```

# Assignment Formats: R Scripts

For your assignments, you will be writing R scripts with comments and code which solve problems and answer questions. You will submit your R scripts on D2L. I will download all the scripts and run them in RStudio on my computer to check your work. Therefore, you will need to follow strict assignment formatting rules. 

## File Names

Each assignment should follow this naming convention:

`full_name_hw#`

For example: `justin_pomeranz_hw1`

You can capitalize words in the file name if you choose, but it should follow the general convention above. 

## File format

* Comment before each problem, and each sub-problem
* Make sure your results print out on `Source with echo`
  + If you're answer is saved in an object, make sure you print out that object afterwards
  + for example, see that `volume` is on the last line of the problem 2 example below

```{r assignment format, eval = FALSE}
# Problem 1

# 1.1
2 - 10

# 1.2
3 * 5

# problem 2

width = 2
height = 3
length = 1.5
volume = width * height * length
volume

```

