---
title: "Homework 4: ggplot"
subtitle: "ENVS475: Experimental Analysis and Design"
date: "Spring 2023"
output: 
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
   warning = FALSE, message = FALSE,
   eval = FALSE,
   fig.align = 'center', fig.retina = 2,
  collapse = TRUE,
  comment = "#>"
)
```

## This assignment will not be graded

# Assignment Format

* Open a new R script
* Save your R script in the `homework/` folder
* Name your script according to: `full_name_hw##.R`
  + For example: `justin_pomeranz_hw01.R`
* Comment before each problem, and each sub-problem
* Make sure your results print out on `Source with echo`
  + If you're answer is saved in an object, make sure you print out that object afterwards
  + See that `volume` is on the last line of the example below

```{r assignment format, eval = FALSE}
# problem 2 example
width = 2
height = 3
length = 1.5
volume = width * height * length
volume
```

# Setup

## Load packages

* Make sure that the `ggplot2` package is downloaded on the machine you're using
  + You can check this by clicking on the `Packages` tab in the *Files, Plots, Packages...* panel and scrolling down
  + If you need to download the package, copy the following code into your Console panel and run it **Do not put it in your homework script*
  + `install.packages("ggplot2")` 
* Copy the following code into the beginning of your homework script to load the `ggplot2` package:
```{r, eval=FALSE}
# load libraries
library(ggplot2)
```

## Data

An experiment in Kenya has been exploring the influence of large herbivores on plants.

Check to see if `uhuru_acacia_lab2.txt` is in your workspace.

If not download it from D2L and put it in your `data/` folder in your Class R project. 

Read the data file into R using the following command:

```{r, eval=FALSE}
acacia <- read.csv("data/uhuru_acacia_lab2.txt", sep="\t", na.strings = c("dead"))
```

# Exercises Homework 04: `ggplot2`

## Problem 1: Basic `ggplot` 

1. Make a scatter plot with `CIRC` on the x axis and `AXIS1` (the maximum canopy
   width) on the y axis. Label the x axis "Circumference" and the y axis "Canopy
   Diameter".

2. The same plot as (1), but with both axes scaled logarithmically (using `scale_x_log10` and `scale_y_log10`).

3. The same plot as (1), but with points colored based on the `ANT` column (the species of ant symbiont living with the acacia)

4. The same plot as (3)), but instead of different colors show different species of ant (values of `ANT`) each in a separate subplot.

5. The same plot as (4) but add a simple model of the data by adding `geom_smooth`.

## Problem 2: Statistical Transformations

1. Make a bar plot of the number of acacia with each mutualist ant species (using the `ANT` column).

2. Make a histogram of the height of acacia (using the `HEIGHT` column). Label the x axis "Height (m)" and the y axis "Number of Acacia".
   
3. Make a plot that shows histograms of both `AXIS1` and `AXIS2`. Due to the way the data is structured you’ll need to add a 2nd geom_histogram() layer that specifies a new aesthetic. To make it possible to see both sets of bars you’ll need to make them transparent with the optional argument alpha = 0.3.
  * i.e.:
```{r, eval = FALSE}
ggplot(data = acacia) +
  geom_histogram(mapping = aes(x = AXIS1)) +
  geom_histogram(mapping = aes(x = AXIS2)) 
```
Also:
* Set the color for `AXIS1` to "red" and `AXIS2` to "black" using the `fill` argument. Label the x axis "Canopy Diameter(m)" and the y axis "Number of Acacia".

4. Use `facet_wrap()` to make the same plot as (3) but with one subplot for each treatment. Set the number of bins in the histogram to 10.

# Problem 3: dplyr + ggplot

### Data
Download the `uhuru_trees.csv` file from D2L and put it in your `data/` folder. 

Load the data into R using the following command:

```{r, eval = FALSE}
trees <- read.csv("data/uhuru_trees.csv")
```

1. Update the `trees` data frame with a new column named `canopy_area` that contains the estimated canopy area calculated as the value in the `AXIS_1` column times the value in the `AXIS_2` column.
Show output of the `trees` data frame with just the `SURVEY`, `YEAR`, `SITE`, and `canopy_area` columns.

2. Make a scatter plot with `canopy_area` on the x axis and `HEIGHT` on the y-axis. Color the points by `TREATMENT` and plot the points for each value in the `SPECIES` column in a separate subplot. Label the x axis "Canopy Area (m)" and the y axis "Height (m)". Make the point size 2.

3. That's a big outlier in the plot from (2). 50 by 50 meters is a little too big for a real Acacia, so filter the data to remove any values for `AXIS_1` and `AXIS_2` that are over 20 and update the data frame. Then remake the graph.

4. Using the data without the outlier (i.e., the data generated in (3)), find out how the abundance of each species has been changing through time.
Use `group_by`, `summarize`, and `n` to make a data frame with `YEAR`, `SPECIES`, and an `abundance` column that has the number of individuals in each species in each year. Print out this data frame.

5. Using the data the data frame generated in (4), make a line plot with points (by using `geom_line` in addition to `geom_point`) with `YEAR` on the x axis and `abundance` on the y axis with one subplot per species. To let you seen each trend clearly let the scale for the y axis vary among plots by adding `scales = "free_y"` as an optional argument to `facet_wrap`.